# Data Wrangling   

<!-- ========================= -->
## How to do data wrangling  
We will use `tidyverse` package to work with data.  

### Load data and package
```{r}
head (iris)
```
```{r}
library(tidyverse)
```
### Select certain rows
```{r}
setosa <- filter(iris, Species == 'setosa')
setosa
```
### Select certain columns
```{r}
select(iris, Sepal.Length, Species)
```

```{r}
select(iris, -Sepal.Length, -Species)
```
### Rename variables
```{r}
rename(iris,  Sepal_Width= Sepal.Width,  Sepal_Length= Sepal.Length )
```
### Sorting in ascending or descending order  
- put a minus in front of a variable for descending order
```{r}
arrange(iris, Petal.Length, -Petal.Width)
 
```
### Transform variables
```{r}
mutate(iris,  newvar= Sepal.Width*10, Petal.Length=Petal.Length/100  )
```
### Working with pipes %>% 
```{r}
iris %>% 
  filter(Species=="setosa") %>% 
  mutate (newvar=Sepal.Width*10) %>% 
  select (-Sepal.Width, -Petal.Width) %>% 
  arrange(-Sepal.Length, newvar)
```
### Pivot wider (long to wide)  
- If no unique identifier row in each group **doesn't work**   

```{r}
iris %>%
pivot_wider(  names_from=Species, values_from= c(Sepal.Length))
```
- Create a unique identifier row for each name and then use pivot_wider 

```{r}
widedata <- iris %>%
  # create groups then assign unique identifier row number in each group
  group_by(Species) %>%
  mutate(row = row_number()) %>%
pivot_wider(names_from=Species, values_from= c(Petal.Length,Sepal.Length,Petal.Width,Sepal.Width))
widedata
```

```{r}
iris %>%
  group_by(Species) %>%
  mutate(row = row_number()) %>%
pivot_wider( names_from=Species, values_from= c(Petal.Length, Petal.Width))
```
### Pivot longer (wide to long)   

```{r}
longdata = pivot_longer(widedata, -
                          c(  "row"   ,                  "Petal.Length_setosa" ,    "Petal.Length_versicolor",
  "Petal.Length_virginica",  "Sepal.Length_setosa"  ,   "Sepal.Length_versicolor",
  "Sepal.Length_virginica" , "Petal.Width_setosa"    ,  "Petal.Width_versicolor" ,
  "Petal.Width_virginica"    )        , names_to="Sepal.Width", values_to="Sepal.Width.value")
longdata
```
- Pivot wider again (long to wide)

```{r}
pivot_wider(longdata, names_from=Sepal.Width, values_from= c(Sepal.Width.value))

```
### Separate columns       
```{r}
separate(iris, Species, into = c("integer","decimal","third"), sep="o")

```
### Recode/relabel data    
```{r}
mutate(iris, Species2 = recode(Species, "setosa"="seto", "versicolor"="versi"))
```
### Combine data sets  
- prepare data sets 
```{r}
data1 <- data.frame(ID = 1:4,                       
                    X1 = c("a1", "a2","a3", "a4"),
                    stringsAsFactors = FALSE)
data2 <- data.frame(ID = 2:5,                       
                    X2 = c("b1", "b2","b3", "b4"),
                    stringsAsFactors = FALSE)
```
- inner join   
```{r}
inner_join(data1, data2, by = "ID")    
```
- left join   
```{r}
left_join(data1, data2, by = "ID")   
```
- right join   
```{r}
right_join(data1, data2, by = "ID")  
```
- full join   
```{r}
full_join(data1, data2, by = "ID") 
```
- keep cases of left data table without in right data table 
```{r}
anti_join(data1, data2, by = "ID")  
```
- keep cases of left data table in right data table 
```{r}
semi_join(data1, data2, by = "ID")  
```
- multiple full join  
```{r}
full_join(data1, data2, by = "ID") %>%              
  full_join(., data2, by = "ID") 
```
- rbind **doesn't work** 
```{r}
# df1 <- data.frame(col1 = LETTERS[1:6],
#                   col2a = c(5:10),
#                   col3a = TRUE)
#   
# df2 <- data.frame(col1 = LETTERS[4:8],
#                   col2b= c(4:8),
#                   col3b = FALSE)
# rbind(df1,df2)
```
- append two data tables by using join and merge       
```{r}
data_frame1 <- data.frame(col1 = c(6:8),
                         col2 = letters[1:3],
                         col3 = c(1,4,NA))
 
data_frame2 <- data.frame(col1 = c(5:6),
                          col5 = letters[7:8])
 
data_frame_merge <- merge(data_frame1, data_frame2,
                          by = 'col1', all = TRUE)
 
print (data_frame_merge)
```
```{r}
full_join(data_frame1,data_frame2, by=c("col1"),)
```



<!-- ======================= -->

## How to do aggregation/ summarization
### Summarization after grouping
```{r echo=TRUE,  message=FALSE}
library(tidyverse)
```

```{r echo=TRUE,  include=FALSE}
library(tidyverse)
```

```{r}
iris %>% 
  group_by(Species) %>% 
  summarize(Support = mean(Sepal.Length)) %>%    # average
  arrange(-Support)                         # sort
```

```{r}
iris %>% 
  group_by(Species) %>% 
summarize(mean_s = mean(Sepal.Width), 
            meas_p = mean(Petal.Length), 
            diff = mean(Sepal.Width-Petal.Length)) %>% 
  arrange(-diff) 
```

```{r}
iris %>% 
  group_by(Species) %>% 
summarize(n = n(), 
            meas_p = mean(Petal.Length), 
            sd = sd(Petal.Length))  
```
### Summarization with upgroup
```{r}
iris %>% 
  ungroup( ) %>% 
summarize(n = n(), 
            meas_p = mean(Petal.Length), 
            sd = sd(Petal.Length))  
```
### Mutate new variables after grouping
```{r}
iris %>% 
  group_by(Species) %>% 
mutate(n = n(), 
            meas_p = mean(Petal.Length), 
            sd = sd(Petal.Length))  
```


```{r}
iris %>% 
  group_by(Species) %>% 
mutate(n = n(), 
            meas_p = mean(Petal.Length, na.rm = T), 
            sd = sd(Petal.Length))  %>% 
  summarize (n_mean = paste ("sample size:",mean(n)), 
            meas_p = mean(Petal.Length), 
            sd = sd(Petal.Length))
```

### Recode and generate new variables, then value label
```{r}
irisifelse <- iris %>% 
mutate(Species2 = ifelse(Species == "setosa", NA, Species))
# relabel values
irisifelse$Species2 <- factor(irisifelse$Species2,labels = c( "versi","virg"))
irisifelse
```
```{r}
str(irisifelse)
```
 
 