# Data Wrangling   

<!-- ========================= -->
## How to do data wrangling  
We will use `tidyverse` package to work with data.  

### Load data and package
```{r}
head (iris)
```
```{r}
library(tidyverse)
```
### Select certain rows
```{r}
setosa <- filter(iris, Species == 'setosa')
setosa
```

`filter the first 5 rows`
```{r}
iris %>% 
  slice(1:5)
```


### Select certain columns
```{r}
select(iris, Sepal.Length, Species)
```

```{r}
select(iris, -Sepal.Length, -Species)
```

`reorder the variables`
```{r}
  iris %>% select(Species, everything()) %>% 
  print()
```

```{r}
# using the direction of vector
  iris %>% select(Species:Sepal.Length ) %>% 
  print()
```

### Rename variables
```{r}
rename(iris,  Sepal_Width= Sepal.Width,  Sepal_Length= Sepal.Length )
```

```{r}
iris %>% 
  rename_with(tolower) %>% head()
```

### Sorting in ascending or descending order  
- put a minus in front of a variable for descending order
```{r}
arrange(iris, Petal.Length, -Petal.Width)
 
```
### Transform variables
```{r}
mutate(iris,  newvar= Sepal.Width*10, Petal.Length=Petal.Length/100  )

# use mutate and ifelse to categorize the varible
```

`rep, sample` 
```{r}
set.seed(123)

drug_trial <- tibble(
  #  each id has three 
  id = rep(1:20, each = 3),
  #  all has 20
  year = rep(0:2, times = 20),
  # sample 20 and each repeat 3  
  age = sample(35:75, 20, TRUE) %>% rep(each = 3),
  # 
  drug = sample(c("Placebo", "Active"), 20, TRUE) %>% 
    rep(each = 3),
  #  
  se_headache = if_else(
    drug == "Placebo", 
    # the possibility that outcome is 1
    sample(0:1, 60, TRUE, c(.95,.05)), 
    sample(0:1, 60, TRUE, c(.10, .90))
  )
  #  
)
head(drug_trial)
```

```{r}
# 
drug_trial %>%
  mutate(complete = c(1))
```

`factor is just for factor (labeling the value; while `rename` labeling the variable) and recode is for different types whose outcome depends on the right side of the equation sign. factor can change the reference group. `.
```{r}
drug_trial=drug_trial %>% 
  mutate(mi_f = factor(se_headache , c(0, 1), c("No", "Yes")))%>%  
  mutate(mi_f_reverse = factor(se_headache , c(1, 0), c("yes", "no")))%>% 
  # change the reference group by using factor
  
  mutate(mi  = recode(se_headache ,  "0"=2, "1"=1))

str(drug_trial)

table (drug_trial$mi_f)
table (drug_trial$mi_f_reverse)
```

```{r}
drug_trial %>% 
  rowwise() %>% 
  mutate(any_se_year = sum(se_headache, year) > 0)
```

`what is the difference of rowwise and group_by`
```{r}
drug_trial %>%  group_by(id) %>% 
  mutate(any_se_year = sum(se_headache, year)  )
```

```{r}

drug_trial %>% 
  rowwise() %>% 
  mutate(any_se_year = sum(se_headache, year)  )
```

`starts_with and end_with`
```{r}
drug_trial_sub <- drug_trial %>% 
  select(id, year, starts_with("se")) %>% 
  print()
```

`keep all combination levels even 0 times`
```{r}
drug_trial %>% 
  filter(year == 0) %>% 
  filter(age < 65) %>% 
  group_by(drug, se_headache, .drop = FALSE) %>% 
  summarise(n = n())
```
`group_by`
```{r}
drug_trial %>% 
  filter(!is.na(age)) %>% 
  group_by(drug) %>% 
  summarise(mean_age = mean(age))
```



### Working with pipes %>% 
```{r}
iris %>% 
  filter(Species=="setosa") %>% 
  mutate (newvar=Sepal.Width*10) %>% 
  select (-Sepal.Width, -Petal.Width) %>% 
  arrange(-Sepal.Length, newvar)
```
### Pivot wider (long to wide)  
- If no unique identifier row in each group **doesn't work**   

```{r}
iris %>%
pivot_wider(  names_from=Species, values_from= c(Sepal.Length))
```
- Create a unique identifier row for each name and then use pivot_wider 

```{r}
widedata <- iris %>%
  # create groups then assign unique identifier row number in each group
  group_by(Species) %>%
  mutate(row = row_number()) %>%
pivot_wider(names_from=Species, values_from= c(Petal.Length,Sepal.Length,Petal.Width,Sepal.Width))
widedata
```

```{r}
iris %>%
  group_by(Species) %>%
  mutate(row = row_number()) %>%
pivot_wider( names_from=Species, values_from= c(Petal.Length, Petal.Width))
```
### Pivot longer (wide to long)   

why long format? there are many analytic techniques that require our data to be in this format, like plotting.
```{r}
longdata = pivot_longer(widedata, -
                          c(  "row"   ,                  "Petal.Length_setosa" ,    "Petal.Length_versicolor",
  "Petal.Length_virginica",  "Sepal.Length_setosa"  ,   "Sepal.Length_versicolor",
  "Sepal.Length_virginica" , "Petal.Width_setosa"    ,  "Petal.Width_versicolor" ,
  "Petal.Width_virginica"    )        , names_to="Sepal.Width", values_to="Sepal.Width.value")
longdata
```


```{r}
babies <- tibble(
  id       = 1001:1008,
  sex      = c("F", "F", "M", "F", "M", "M", "M", "F"),
  weight_3  = c(9, 11, 17, 16, 11, 17, 16, 15),
  weight_6  = c(13, 16, 20, 18, 15, 21, 17, 16),
  weight_9  = c(16, 17, 23, 21, 16, 25, 19, 18),
  weight_12 = c(17, 20, 24, 22, 18, 26, 21, 19)
) %>% 
  print()
```

`select which cols`
```{r}
babies %>% 
  pivot_longer(
    cols         = starts_with("weight"),  #which cols
    names_to     = "months",
    names_prefix = "weight_",  #remove prefix 
    values_to    = "weight"
  )
```

`using names_prefix argument`

“it is a regular expression used to remove matching text from the start of each variable name.”
```{r}
babies %>% 
  pivot_longer(
    cols         = starts_with("weight"),
    names_to     = "months",
    names_prefix = "\\w+_"
  )
```




`using default argument`
```{r}
babies %>% 
  pivot_longer(
    cols = starts_with("weight")
  )
```

```{r}
set.seed(123)
babies <- tibble(
  id       = 1001:1008,
  sex      = c("F", "F", "M", "F", "M", "M", "M", "F"),
  weight_3  = c(9, 11, 17, 16, 11, 17, 16, 15),
  weight_6  = c(13, 16, 20, 18, 15, 21, 17, 16),
  weight_9  = c(16, 17, 23, 21, 16, 25, 19, 18),
  weight_12 = c(17, 20, 24, 22, 18, 26, 21, 19),
  length_3  = c(17, 19, 23, 20, 18, 22, 21, 18),
  length_6  = round(length_3 + rnorm(8, 2, 1)),
  length_9  = round(length_6 + rnorm(8, 2, 1)),
  length_12 = round(length_9 + rnorm(8, 2, 1)),
) %>% 
  print()
```
- pivoting multiple sets of cols
```{r}
babies %>% 
  pivot_longer(
    cols      = c(-id, -sex),
    names_to  = c(".value", "months"),
    names_sep = "_"
  )
```

```{r}
babies %>% 
  pivot_longer(
    cols      = c(-id, -sex),
    names_to  = "months",
    values_to = "weight"
  ) %>% 
  print()
```

`using names_sep and .value argument `
```{r}
babies %>% 
  pivot_longer(
    cols      = c(-id, -sex),
    names_to  = c("measure", "months"),
    names_sep = "_"
  ) 
```

`.value tells pivot_longer() to create a new column for each unique character string that is in front of the underscore`

```{r}
babies %>% 
  pivot_longer(
    cols      = c(-id, -sex),
    names_to  = c( ".value","months" ), 
    names_sep = "_"
  )  
```

 

```{r}
babies %>% 
  pivot_longer(
    cols      = c(-id, -sex),
    names_to  = c( "months", ".value"),
    names_sep = "_"
  )
```



- Pivot wider again (long to wide)

`pivoting to multiple sets of cols`
```{r}
babies %>% 
  pivot_longer(
    cols      = c(-id, -sex),
    names_to  = c( ".value","months" ), 
    names_sep = "_"
  ) %>% 
  pivot_wider( names_from=months, values_from= c(weight, length ))
```

```{r}
pivot_wider(longdata, names_from=Sepal.Width, values_from= c(Sepal.Width.value))

```
### Separate columns       
```{r}
separate(iris, Species, into = c("integer","decimal","third"), sep="o")

```
### Recode/relabel data    
```{r}
mutate(iris, Species2 = recode(Species, "setosa"="seto", "versicolor"="versi"))
```

### deduplication
#### Complete duplicate row
```{r}
df <- tribble(
  ~id, ~day, ~x,
    1, 1, 1,
    1, 2, 11,
    2, 1, 12,
    2, 2, 13,
    2, 2, 14,
    3, 1, 12,
    3, 1, 12,
    3, 2, 13,
    4, 1, 13,
    5, 1, 10,
    5, 2, 11,
    5, 1, 10
)  

df %>% 
  distinct()
```

`mark duplication`
```{r}
df %>% 
  mutate(dup = duplicated(df)) 
  
```

```{r}
df %>% 
  group_by_all() %>% 
  mutate(
    n_row = row_number(),
    dup   = n_row > 1
  )
```

#### Partial duplicate rows
```{r}
df %>% 
  distinct(id, day, .keep_all = TRUE)
```

```{r}
df %>% 
  group_by(id, day) %>% 
  mutate(
    count = row_number(), # Counts rows by group
    dup   = count > 1     # TRUE if there is more than one row per group
  )
```

`select duplicates`
```{r}
df %>% 
  mutate(dup = duplicated(.) | duplicated(., fromLast = TRUE)) %>% 
  filter(dup==T)
```

### Combine data sets  
- prepare data sets 
```{r}
data1 <- data.frame(ID = 1:4,                       
                    X1 = c("a1", "a2","a3", "a4"),
                    stringsAsFactors = FALSE)
data2 <- data.frame(ID = 2:5,                       
                    X2 = c("b1", "b2","b3", "b4"),
                    stringsAsFactors = FALSE)
```
- inner join   
```{r}
inner_join(data1, data2, by = "ID")    
```
- left join   

`there is not ID 1 in data2, thence, x2 is missing. `
```{r}
left_join(data1, data2, by = "ID")   
```

`differing key col names`
```{r}
left_join(data1, data2, by = c("ID"="ID"))  
```

`One-to-many relationship` has the same statement  
```{r}
left_join(data1, data2, by = c("ID" ))  
```

`multiple data frames`
```{r}
data1 %>% 
  left_join(data2, by = "ID") %>% 
  left_join(data2, by = "ID")
```

`multiple key values`
```{r}

# demographics %>%  
#   left_join(ultra, by = c("id", "visit"))
```

- right join   
```{r}
right_join(data1, data2, by = "ID")  
```
- full join   
```{r}
full_join(data1, data2, by = "ID") 
```
- keep cases of left data table without in right data table 
```{r}
anti_join(data1, data2, by = "ID")  
```
- keep cases of left data table in right data table 
```{r}
semi_join(data1, data2, by = "ID")  
```
- multiple full join  
```{r}
full_join(data1, data2, by = "ID") %>%              
  full_join(., data2, by = "ID") 
```

 
- append two data tables by using join and merge       
```{r}
data_frame1 <- data.frame(col1 = c(6:8),
                         col2 = letters[1:3],
                         col3 = c(1,4,NA))
 
data_frame2 <- data.frame(col1 = c(5:6),
                          col5 = letters[7:8])
 
data_frame_merge <- merge(data_frame1, data_frame2,
                          by = 'col1', all = TRUE)
 
print (data_frame_merge)
```

```{r}
full_join(data_frame1,data_frame2, by=c("col1"),)
```

- adding rows using `bind_ function`

rbind **doesn't work** 

`bind_ function conbines the data frames `based on column names`; having our columns in a different order isn’t a problem.` 
```{r}
df1 <- data.frame(col1 = LETTERS[1:6],
                  col2a = c(5:10),
                  col3a = TRUE)

df2 <- data.frame(col1 = LETTERS[4:8],
                  col2b= c(4:8),
                  col3b = FALSE)
# rbind(df1,df2)

df1 %>% 
  bind_rows(df2)
```

`rename the differing cols`
```{r}
df1 %>% 
  bind_rows(df2 %>% 
   rename(col2a=col2b,
          col3a=col3b)  )
```


combining more than 2 data frames 
```{r}
df1 %>% 
  bind_rows(df2,df2)
```



### Working with character strings
```{r}
library(stringr)  
library(readxl)
```
- Look at the values   
```{r}
ehr <- read_excel("C:\\Users\\hed2\\Downloads\\mybook2\\mybook2/excel.xlsx")
ehr %>% 
  arrange(sex) %>% 
  pull(sex)
```
- Coerce to lowercase
```{r}
ehr %>% 
  arrange(sex) %>% 
  pull(sex) %>% 
  str_to_lower()
```
- Coerce to upper case
```{r}
ehr %>% 
  arrange(sex) %>% 
  pull(sex) %>% 
  str_to_upper()
```
- Title case
```{r}
ehr %>% 
  arrange(sex) %>% 
  pull(sex) %>% 
  str_to_title()
```
- Sentence case
```{r}
ehr %>% 
  arrange(sex) %>% 
  pull(sex) %>% 
  str_to_sentence()
```
- Trim white space from the beginning and end 
```{r}
str_trim("Ryan Edwards  ")
```
- Remove the comma
```{r}
str_replace(
  string      = "weston fox,", 
  pattern     = ",",
  replacement = ""
)
```

```{r}
ehr %>% 
  mutate(sex = str_replace(sex, ",", ""))
```

- Separate values into component parts 

regular express, `^\\w+ look for one or more consecutive word characters at the start of the character string and extract them.`

```{r}
str_extract("zariah hernandez", "^\\w+")
str_extract("zariah hernandez", "\\w+$")
```

```{r}
stringr::str_match("mean(weight_3)", "(\\()(\\w+)(\\_)(\\d+)(\\))")
```

- Detect a special string to category
`index a special string`
```{r}
ehr %>% 
  mutate(
    man   = str_detect(sex, "F"),
    woman = str_detect(sex, "M"),
    all   = str_detect(sex, "ale")
  ) %>% 
  mutate(man_dummy = as.numeric(man))
```


```{r}
ehr %>% 
  mutate(
    dummy   =   ifelse(str_detect(sex, "F"),0,1)   
  )  
```

### Conditional operations

- Testing multiple conditions simultaneously
```{r}
blood_pressure <- tibble(
  id     = 1:10,
  sysbp  = c(152, 120, 119, 123, 135, 83, 191, 147, 209, 166),
  diasbp = c(78, 60, 88, 76, 85, 54, 116, 95, 100, 106)
) %>% 
  print()
```

```{r}
blood_pressure %>% 
  mutate(bp = if_else(sysbp < 120 & diasbp < 80, "Normal", "Not Normal"))
```

- Using case_when function
```{r}
blood_pressure %>% 
  mutate(
    bp = case_when(
      sysbp < 120 & diasbp < 80                               ~ "Normal",
      sysbp >= 120 & sysbp < 130 & diasbp < 80                ~ "Elevated",
      sysbp >= 130 & sysbp < 140 | diasbp >= 80 & diasbp < 90 ~ "Hypertension Stage 1",
      sysbp >= 140 | diasbp >= 90                             ~ "Hypertension Stage 2"
    )
  )
```
 
- Recoding variables
```{r}
blood_pressure %>% 
  mutate(
    bp= case_when(
      sysbp < 120 & diasbp < 80                               ~ 1,
      sysbp >= 120 & sysbp < 130 & diasbp < 80                ~ 2,
      sysbp >= 130 & sysbp < 140 | diasbp >= 80 & diasbp < 90 ~ 3,
      sysbp >= 140 | diasbp >= 90                             ~ 4
    )
  ,
 bp_f = factor( bp,
      labels = c(
        "normal", "elevated", "stage 1",
        "stage 2")
      )
  )
```

- Recoding missing

`using NA_real instead of NA.`
```{r}
demographics <- tibble(
    race     = c(1, 2, 1, 4, 7, 1, 2, 9, 1, 3),
    hispanic = c(7, 0, 1, 0, 1, 0, 1, 9, 0, 1)
  )

demographics %>% 
  mutate(
    # Recode 7 and 9 to missing
    race_recode = if_else(race == 7 | race == 9, NA_real_, race),
    hispanic_recode = if_else(hispanic == 7 | hispanic == 9, NA_real_, hispanic)
  ) 
```


`check whether or not `
```{r}
blood_pressure %>% 
  mutate(
    match = sysbp == sysbp,
    n_match = if_else(is.na(match), FALSE, match)
  )
```
 

<!-- ======================= -->

## How to do aggregation/ summarization
### Summarization after grouping
```{r echo=TRUE,  message=FALSE}
library(tidyverse)
```

```{r echo=TRUE,  include=FALSE}
library(tidyverse)
```

```{r}
iris %>% 
  group_by(Species) %>% 
  summarize(Support = mean(Sepal.Length)) %>%    # average
  arrange(-Support)                         # sort
```

```{r}
iris %>% 
  group_by(Species) %>% 
summarize(mean_s = mean(Sepal.Width), 
            meas_p = mean(Petal.Length), 
            diff = mean(Sepal.Width-Petal.Length)) %>% 
  arrange(-diff) 
```

```{r}
iris %>% 
  group_by(Species) %>% 
summarize(n = n(), 
            meas_p = mean(Petal.Length), 
            sd = sd(Petal.Length))  
```
### Summarization with upgroup
```{r}
iris %>% 
  ungroup( ) %>% 
summarize(n = n(), 
            meas_p = mean(Petal.Length), 
            sd = sd(Petal.Length))  
```
### Mutate new variables after grouping
```{r}
iris %>% 
  group_by(Species) %>% 
mutate(n = n(), 
            meas_p = mean(Petal.Length), 
            sd = sd(Petal.Length))  
```


```{r}
iris %>% 
  group_by(Species) %>% 
mutate(n = n(), 
            meas_p = mean(Petal.Length, na.rm = T), 
            sd = sd(Petal.Length))  %>% 
  summarize (n_mean = paste ("sample size:",mean(n)), 
            meas_p = mean(Petal.Length), 
            sd = sd(Petal.Length))
```

### Recode and generate new variables, then value label
```{r}
irisifelse <- iris %>% 
mutate(Species2 = ifelse(Species == "setosa", NA, Species))
# relabel values
irisifelse$Species2 <- factor(irisifelse$Species2,labels = c( "versi","virg"))
irisifelse
```
```{r}
str(irisifelse)
```
 
## How to creat table 1 with test

[see here](https://rpubs.com/Daniel_He/1041833)

