# Data visualization

<!-- ================ -->
## Data visualization introduction

### Summarization 
```{r}
library(tidyverse)
library(dplyr)
mtcars %>%
  mutate( kml = mpg * 0.42) %>%
  group_by(cyl) %>%
   summarise(avg_US = mean(mpg), avg_metric = mean(kml))
```

```{r}
mpg %>% 
  group_by(manufacturer, year) %>% 
  summarise_at(vars(cty, hwy), mean)
```
- change layout
```{r}
mpg %>% count(class, year)%>% 
  spread(class, n) 
```
- change all characters into factors
```{r}
mpg <- mpg %>% 
  mutate_if(is.character, as.factor) #if a column is a character, change to a factor
```
- wide to long data
```{r}
mpg1 <- mpg %>% 
  gather("key", "value", cty, hwy)
```

- convert wide data to long data using `pivot_longer `
```{r}
## Your code here. Naming choices for 1 and 2 are yours
dta <- mpg %>% 
  pivot_longer(cty:hwy, names_to = "var", values_to = "value") %>%  # Both of those are 
  # value label
   mutate(var = ifelse( var == 'cty', 'city','highway'))

ggplot(dta, aes(x = displ, y = value)) + 
  geom_point(aes(color = var))  +
  geom_smooth(aes(color = var), se = F)
```

- explore distribution  
```{r}
library(DataExplorer)
library(psych)
library(naniar)
plot_histogram(riskfactors)
```

- explore relationship/correlation
```{r}
library(psych)
pairs.panels(riskfactors[,1:10])
```

- create a individual theme
```{r}
my_theme <- function(){
  theme_bw() + 
 
  theme(axis.title = element_text(size=16),
  axis.text = element_text(size=14),
  text = element_text(size = 14))
}
```
### Explore missing values
```{r}
# install.packages("naniar")
library(naniar)
# head(riskfactors)
riskfactors <- riskfactors
```

```{r}
gg_miss_upset(riskfactors,nsets=10)
```

```{r}
# install.packages("DataExplorer")

plot_missing(riskfactors)
```

```{r}
# take a quick look at the data types of each column
visdat::vis_dat(riskfactors)
```
 



### Add statistical test 
```{r ,warning=F}
library(ggpubr)
plt <- ggplot(  data=mpg,
  mapping= aes(x =  as.factor(year),
               y = cty,
               color = as.factor(year) ) )+
  geom_boxplot() +
  geom_jitter(width=0.1)+
  labs(x = 'Year',
       y = "City mpg") +
  my_theme()+ facet_wrap( ~ manufacturer,nrow = 2)  
  
# add statistical test
  my_comparisons <- list(c('1999','2008'))
plt + stat_compare_means() +
stat_compare_means(comparisons = my_comparisons)
```


### Add texts to dots 
```{r}
USArrests <- USArrests %>% rownames_to_column('State')

ggplot(USArrests, aes(
x=UrbanPop,y=Murder))+
geom_point() +
  labs(x = "Percent of population that is urban", 
       y = "Murder arrests (per 100,000)",
       caption =  "McNeil (1997). Interactive Data Analysis")+
  geom_text(aes(label=State),size=3)
```

### Set the legend
```{r}
ggplot(iris, aes(x=  Sepal.Length , fill= as.factor( Species)) ) +  #whole plot's option
  
  geom_histogram(aes(y=..density..),alpha=0.5, position="identity" , bins = 50)+
  geom_density(aes(linetype=as.factor(Species)),alpha=.1 )+    #aesthetic's option
  
  scale_fill_manual( name = "Groups",values = c("grey", "black", "skyblue"),labels = c("setosa", "versicolor" , "virginica" ))+  
  scale_linetype_manual( name = "Groups" ,values = c(1,3,5),labels = c("setosa", "versicolor" , "virginica") )+       # common legend 
  
  labs(x = "Sepal.Length", 
       y = "Density", 
       title = "") 
```

### Create a panel of plots

- combine multiple plots into one
```{r}
p1=ggplot(data=riskfactors,aes(x=age))+
  geom_histogram(bins = 30 )

p2=ggplot(data=riskfactors,aes(x=sex))+
  geom_bar (aes(x=sex) )

p3=ggplot(riskfactors,aes(x = education, y = bmi))+
  geom_boxplot (  )

p4=ggplot(riskfactors, aes(x = marital )) + 
  geom_bar(aes(group = education, y = (..count..)/sum(..count..),fill = education)) + 
          scale_y_continuous(labels=scales::percent) 

# install.packages("ggpubr")
library(ggpubr)
ggarrange(p1, p2, p3, p4, ncol = 2, nrow=2)
```



### Plots in regression
- create linear regression model
```{r}
data("Boston", package = "MASS")
linear_reg <- glm(medv ~ ., data=Boston , family = gaussian())
summary(linear_reg)
```
- summary
```{r}
knitr::kable(broom::tidy(linear_reg))
```

- create logistical regression 
```{r}
# load the Pima Indians dataset from the mlbench dataset
library(mlbench)
data(PimaIndiansDiabetes) 
# rename dataset to have shorter name because lazy
diabetes <- PimaIndiansDiabetes

logistic_reg <- glm(diabetes ~ ., data=diabetes, family = binomial)

summary(logistic_reg)

```
- summary

```{r}
knitr::kable(broom::tidy(logistic_reg))
```


#### Create forest plots for `coefficients` or `OR`
```{r}
library(sjPlot)
plot_model(linear_reg, show.values = TRUE, value.offset = 0.5)
```

```{r}
plot_model(logistic_reg, show.values = TRUE, value.offset = .5, vline.color = "black")
```
**another way**
```{r}
library(finalfit) 

explanatory = c(  "crim"  ,  "zn"   ,   "indus"  ,   "nox"   ,  "rm"   ,   "age" ,    "dis"  ,   "rad"   ,  "tax"   ,"ptratio" ,"black"  , "lstat" )

dependent = "medv"


Boston %>%
 coefficient_plot(dependent, explanatory, table_text_size=3, 
                  title_text_size=12,
                   plot_opts=list(xlab("Beta, 95% CI"), 
                                  theme(axis.title = element_text(size=12))))
```

```{r}
 
library(finalfit) 

explanatory = c(  "pregnant", "glucose" , "pressure", "triceps"  ,"insulin" , "mass"   ,  "pedigree", "age"  )

dependent = "diabetes"

diabetes %>%
 or_plot(dependent, explanatory, table_text_size=3, 
                  title_text_size=12,
                   plot_opts=list(xlab("OR, 95% CI"), 
                                  theme(axis.title = element_text(size=12))))
 
```

- qq plot
```{r}
ggqqplot( (Boston$medv))
```

<!-- ======================== -->

- Loading data set
```{r}
library(printr)
library(tidyverse)
head(iris)
```
## Scatter plot
### Create a empty canvas
- then create `aesthetic mapping`  
- tell the function which dataset and variables to use
```{r}
ggplot(data = iris,        # which data set? canvas? 
       aes(x=Sepal.Length , y=Petal.Length   ))  # which variables as aesthetics? x and y are mapped to columns of the data; different geoms can have different aesthetics (different variables). 
```

 
### Add a layer/geom of `points` to the canvas
```{r}
ggplot(data = iris,
       mapping = aes(x=Sepal.Length , y=Petal.Length  )) + 
  geom_point()   # adding the geometrical representation 
```
 
```{r,eval=F}
# same plot as above
ggplot(data = iris) + 
  geom_point(  aes(x=Sepal.Length , y=Petal.Length  )) 
```
### Add another aesthetic
- add a curve/straight line to fit these points
- geom provides the aesthetic to ggplot
```{r}
# Loess curve
ggplot(data = iris, 
       mapping = aes(x=Sepal.Length , y=Petal.Length  )) + 
  geom_point() +
  geom_smooth() 

```

```{r}
# Linear regression line
ggplot(data = iris, 
       mapping = aes(x=Sepal.Length , y=Petal.Length)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

### Add other aesthetic
- set other aesthetics `colour, alpha (transparency), and size of points`

```{r}
ggplot(data = iris) + 
  geom_point(aes(x=Sepal.Length , y=Petal.Length, size = Sepal.Width     ), 
             alpha = .5, 
             colour = "red")
```

 

```{r}
ggplot(data = iris) + 
  geom_point(aes(x=Sepal.Length , y=Petal.Length, size = Sepal.Width , colour=Species), #white is a variable here 
             alpha=.9)
```

 
- categorize `Petal.Width` then map colour to this new variable
```{r}
iris <- iris %>% 
  mutate(growth = ifelse(Petal.Width   > 1.5, "Wide", "Normal"))

ggplot(data=iris) + 
  geom_point(aes(x=Sepal.Length , y=Petal.Length, size = Sepal.Width , colour=growth), 
             alpha=.9)
```

 
## Bar chart
```{r}
ggplot(data = iris) + 
  geom_bar(aes(x = growth))
```

- bar chart after `group_by` 
- then use `stat='identity'` 
```{r}
library(dplyr)
results  <- iris %>% 
  group_by(Species, growth) %>% 
   summarise(Sepal.Length.mean=mean (Sepal.Length ))

 gop <- results  %>% 
  filter(Species != "setosa_null"  )
 gop
```
- though meaningless below until line chart (just use the mean as the sum for demonstration)
```{r}
# We can also store parts of a plot in an object
plot1 <- ggplot(gop) + 
  geom_bar(aes(x=growth , y=Sepal.Length.mean), 
           
           stat='identity')
plot1
```
### Add some options for the whole ggplot rather than layers
- `switch` the x and y axes 
```{r}
plot1 + 
  coord_flip()
```

- `reorder` x categories (`-`means descending)
```{r}
ggplot( gop) + 
  geom_bar(aes(x=reorder(growth, -Sepal.Length.mean), y=Sepal.Length.mean, fill=growth), 
           stat='identity') + 
  coord_flip()
```

- add x axis `label` and a `theme` 
```{r}
ggplot(gop) + 
  geom_bar(aes(x=reorder(growth, -Sepal.Length.mean), y=Sepal.Length.mean, fill=growth), 
           stat='identity') + 
  coord_flip() + 
  xlab("Growth categories") + 
  guides(fill=F) +
  theme_minimal()
```

- set theme
```{r}
library(ggthemes)
ggplot(data = iris) + 
  geom_bar(aes(x = growth)) + 
  theme_economist()
```

### Grouped bar chart 
-bar chart with different panels
```{r}
ggplot(mpg, aes(x = class)) + 
  geom_bar() + 
  facet_wrap( ~ year)
```

- actual number (groups are stacked by default)
```{r}
ggplot(gop) + 
  geom_bar(aes(x=growth, y=Sepal.Length.mean, fill=Species)
           , stat='identity'
           )
```

```{r}
ggplot(mpg, aes(x = class )) + 
  geom_bar(aes(group = year, fill = year), position = "stack")
```

- percentage
```{r}
ggplot(gop) + 
  geom_bar(aes(x=growth, y=Sepal.Length.mean, fill=Species), 
           stat='identity', 
           position='fill')
```
 
- groups are dodge with `actual number `
```{r}
ggplot(gop) + 
  geom_bar(aes(x=growth, y=Sepal.Length.mean, fill=Species), 
           stat='identity', 
           position='dodge')

```
- groups are dodge with `percentage`
```{r}
gop2 <- gop %>% 
  group_by(growth ) %>% 
  mutate(Sepal.Length.prop=Sepal.Length.mean/sum(Sepal.Length.mean))

ggplot(gop2) + 
  geom_bar(aes(x=growth, y=Sepal.Length.prop, fill=Species), 
           stat='identity', 
           position='dodge') + 
  ylab("Votes (%)")
```

 
## Line charts
 
```{r}
ggplot(iris) + 
  geom_line(aes(x=Sepal.Length , y=Petal.Length))
```
 
### Grouped by `colour variable`
```{r}
ggplot(iris) + 
  geom_line(aes(x=Sepal.Length , y=Petal.Length, colour = Species))
```

 
- grouped by state then set how many rows or columns
```{r}
ggplot(iris) + 
  geom_line(aes(x=Sepal.Length , y=Petal.Length) ) + 
  
  facet_wrap(~Species, nrow = 1) +      #set how many rows
  coord_flip()  
```

### Multiple aesthetics
```{r}
iris <- iris %>% 
  mutate(growth = ifelse(Petal.Width   > 1.5, "Wide", "Normal"))

ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) + 
  geom_line(size=2,color="purple")+  
  # number format
  scale_x_log10(labels = scales::label_number())+
  geom_point( aes(size =  Sepal.Length,colour = as.factor(growth)),show.legend = F)+
  facet_wrap(~ Species) 

```


## ggplot2 parameters
For detail, please read this [article](http://www.sthda.com/english/wiki/be-awesome-in-ggplot2-a-practical-guide-to-be-highly-effective-r-software-and-data-visualization#change-colors-automatically-and-manually) and this [one](https://ggplot2-book.org/). 

<!-- https://bookdown.org/wangminjie/R4DS/tidyverse-ggplot2-scales.html -->

```{r, echo=FALSE}
packages <- c(
  'ggplot2',
  'ggrepel',
  'patchwork',
  'ggraph',
  'dplyr',
  'gganimate',
  'gifski',
  'ggforce',
  'ggthemes',
  'maps',
  'sf',
  'tidyr',
  'concaveman',
  'remotes'
)
# Uncomment the next lines (remove the first #) to install the packages
# install.packages(packages)
# remotes::install_github("hrbrmstr/hrbrthemes")
# Load standard packages
library(ggplot2)
```


```{r}
library(datasauRus)
  ggplot( )+
    geom_point(data=datasaurus_dozen[datasaurus_dozen$dataset=="dino",], aes(x = x, y = y),color= "#7CAE00" ) +
    theme_void()+
    theme(legend.position = "none") 
 
```


### Components of plot
- Components of plot
  - Data: is a data frame
  - Aesthetics: is used to indicate x and y variables and to control the color, the size or the shape ...
  - Geometry: the type of graphics (bar plot, line plot, scatter plot...)
- adjust parameters
  - adjust legend using `guide_`  
  - adjust color, size, and shape using `scale_`, `guide_` can further adjust `scale_`
  - adjust panel, background, axis (font, color, size, angle), title, legend (position), caption using theme
- types of plots
  - geom_boxplot(): Box plot
  - geom_violin(): Violin plot
  - geom_dotplot(): Dot plot
  - geom_jitter(): Jitter charts
  - geom_line(): Line plot
  - geom_bar(): Bar plot
  - geom_errorbar(): Error bars
  - geom_point(): Scatter plot
  - geom_smooth(): Add smoothed curve
  - geom_quantile(): Add quantile lines  
  - geom_text(): Textual annotations
  - geom_density(): Create a smooth density curve
  - geom_histogram(): Histogram

`example`  

```{r}
data("faithful")
# Basic scatterplot
ggplot(data = faithful, 
       mapping = aes(x = eruptions, y = waiting)) + 
  geom_point()+
  stat_density_2d(aes(fill = ..level..), geom="polygon")   
  
# Data and mapping can be given both as global (in ggplot()) or per layer
# ggplot() + 
#   geom_point(mapping = aes(x = eruptions, y = waiting),
#              data = faithful)
```

### Create main title, axis labels, caption 

pay attention whether argument is factor or continuous.
```{r}
ggplot(data = faithful, 
       mapping = aes(x = eruptions, y = waiting)) + 
  geom_point()+
  
  labs(title = "Number of xxx",
       caption = "source: http://xxx",
       x = "Eruptions" ,
       y = "Waiting time" ) +
# customize title, axis, caption  
  theme(
plot.title = element_text(color="red", size=14, face="bold.italic"),
plot.caption = element_text(color="red", size=10, face="italic"),
axis.title.x = element_text(color="blue", size=14, face="bold"),
axis.title.y = element_text(color="#993333", size=14, face="bold")
)+
  
# hide main title  
theme(plot.title = element_blank() )
```

### Create legend title, position  

```{r}
p <- ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting, colour = eruptions < 3))+
  labs(color="Numbers of eruptions") +
  
  theme(legend.position = "left")+
# Change the appearance of legend title and labels
  theme(legend.title = element_text(colour="blue"),
          legend.text = element_text(colour="red"))+
# Change legend box background color
  theme(legend.background = element_rect(fill=NULL))
print(p)
```

- `customize legends using scale functions`
```{r}
# how to change order of legend?  
# Set legend title and labels
p+ scale_color_discrete(name = "Numbers of eruptions change", labels = c("F", "T" ))
```
 
- customize legend

`guide_colorbar(): continuous colors`

`guide_legend(): discrete values (shapes, colors)`
```{r}
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting, size = eruptions < 3))+
  guides(size = guide_legend(
    # legend title
    title = "title is too low",
    title.position = "bottom",
    title.vjust = -5,
    # legend label
    label.position = "left",
    label.hjust = 1,
    label.theme = element_text(size = 15,
                               face = "italic",
                               colour = "red",
                               angle = 0),
    # label reverse
    reverse = TRUE,
    
    # width of bin
    keywidth = 1,
    
    ncol = 4
    
   )
  )
```

 


- delete a legend
```{r}
  ggplot(mpg, aes(x = displ, y = hwy, color = class, size = cyl)) +
  geom_point() +
  guides(
    color = guide_legend("type"),      # keep
    size = "none"                      # remove
  )
```

- combine two legends when they use the same variable (mapping)
```{r}
  ggplot(mpg, aes(x = displ, y = hwy, color = cyl, size = cyl)) +
  geom_point() +
  scale_color_viridis_c() +
  
  guides(
    color = guide_legend("title"),
    size = guide_legend("title")
  )

# guide = "legend"
```

### Change plot colors 
`set color into aes()` 

```{r}
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting, colour = eruptions < 3))
```

`set color outside of aes()`

```{r}
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting),
             colour = 'steelblue')
```

- Colour the histogram with `color` and `fill `

```{r}
ggplot(faithful) + 
  geom_histogram(aes(x = eruptions,color=eruptions < 3, fill=eruptions < 4))
```
  

- Colour the histogram by `waiting` and changing `position`

```{r}
ggplot(faithful) + 
  geom_histogram(aes(x = eruptions,color=waiting>60), position = 'dodge')
```

```{r}
ggplot(faithful) + 
  geom_histogram(aes(x = eruptions,color=waiting>60), position = 'identity')
```

For fill and stack position, please see `position` section.  


- `change colors manually using scale`
```{r,message=FALSE}
ggplot(faithful) + 
  geom_histogram(aes(x = eruptions,color=eruptions < 3, fill=eruptions < 4))+
  # Box plot
  scale_fill_manual(values=c("#999999", "#E69F00"))
# Scatter plot
  scale_color_manual(values=c( "#E69F00", "#56B4E9"))
```

- `using scale brewer automatically`   
```{r}
ggplot(faithful) + 
  geom_histogram(aes(x = eruptions,color=eruptions < 3, fill=eruptions < 4))+
  # Box plot
  scale_fill_brewer(palette="Dark2")
# Scatter plot
  scale_color_brewer(palette="Set1")
  # using guide to change the color of legend key
  
```

- `using gray colors using scale`
```{r}
# 
p + scale_fill_grey()  #no fill element
# 
p + scale_color_grey()  
```

- `Gradient or continuous colors (can set the middle point as `white`)`
```{r}
# Color by cty values
sp2<-ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = cty))
sp2
# Change the low and high colors
# Sequential color scheme
sp2+scale_color_gradient(low="blue", high="red")
# Diverging color scheme
mid<-mean(mpg$cty)
sp2+scale_color_gradient2(midpoint=mid, low="blue", mid="white",
                          high="red"  )
```


### Change points shapes, transparent and size

make the points larger and slightly transparent.
 
```{r}
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting, shape= eruptions < 3, size=eruptions), color="steelblue", alpha=0.5)

```
 
```{r}
# hwo to reverse order of legend size 
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting, shape= eruptions < 3, size=eruptions), color="steelblue", alpha=0.5)+

  scale_shape_manual(values=c(10, 23 ))+
 
  theme(legend.position="top")
```


### Change bars position
```{r}
p <- ggplot(mpg, aes(fl, fill = drv))

p1 <- p + geom_bar ()

p2 <- p + geom_bar(position = "dodge")
 
p3 <-p + geom_bar(position = "fill")
 
p4 <-p + geom_bar(position = "stack")

library(gridExtra)
grid.arrange(p1, p2, p3,p4, ncol=2)
 
```


### Add text annotations
```{r}
ggplot(data=mpg[(1:100), ], aes(x = displ, y = hwy)) +
  geom_point(aes(color = cty))+
  
  geom_text(aes(label = manufacturer ),
               size = 2, vjust = -1)  #vjust is site not direction
```

### Add a line that (separates points)   

```{r}
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting))+
  
  geom_abline(slope=-13,intercept = 100,color="red", linetype = "dashed")+
  # Add horizontal line at y = 2O; change line type and color
  geom_hline(yintercept=20, linetype="dotted", color = "red")+
# Add vertical line at x = 3; change line type, color and size
  geom_vline(xintercept = 3, color = "blue", size=1.5)
# Add regression line
```

`add segment and arrow`
```{r}
ggplot(mpg, aes(x = displ, y = hwy )) +
  geom_point() +
# Add horizontal line segment
  geom_segment(aes(x = 2, y = 15, xend = 3, yend = 15, size=3, color="red")) +
  geom_segment(aes(x = 3, y = 33, xend = 2.5 , yend = 30),
                  arrow = arrow(length = unit(0.5, "cm")))

```

`fitted curve`
```{r}
ggplot(data=mpg[mpg$fl!="c",], aes(x = displ, y = hwy)) +
  geom_point( ) +
  geom_smooth(color="red") #fitted curve
  
```

```{r}
ggplot(data=mpg[mpg$fl!="c",], aes(x = displ, y = hwy)) +
  geom_point( ) + 
  geom_quantile() +
  geom_rug()+
  theme_minimal()
  
```

`customize line`
```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = fl, linetype = fl, size=fl)) +
  geom_point() +
  geom_line(aes( ) )+
  labs(color="What is fl")+
  
# customize linetype, color, size  
  scale_linetype_manual(values=c("twodash", "dotted","twodash", "dotted","twodash"))+
  scale_color_manual(name = "continents",breaks = c("c", "d", "e", "Europe", "Oceania"),labels = c("africa", "americas", "asia", "europe", "oceania"),  values=c('#999999','#E69F00','#999999','#E69F00','#999999')   )+ #using breaks define three labels   
  scale_size_manual(values=seq(1,4, 0.2))+
  theme(legend.position="top") +
  
  guides(color="legend")
```


 

### Using scale_ function
all mappings have associated scales even if not specified. 

- uisng `scale_colour_brewer`. 

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class)) + 
  scale_colour_brewer(type = 'qual')
```



`RColorBrewer`
```{r}
RColorBrewer::display.brewer.all()
```

- using different palettes 

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class)) + 
  scale_colour_brewer (palette = 'Paired')
```
 
- showing `cyl` with size 
```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour =  class, size=cyl)) + 
  scale_colour_brewer(palette  = 'Set1' ) +
  scale_size (breaks = c(4,6))   

```

<!-- - **Can use `breaks` to select the legend you want** -->
<!-- ```{r} -->
<!-- # how to remove a legend? -->
<!-- ggplot(mpg) +  -->
<!--   geom_point(aes(x = displ, y = hwy, colour =  class, size=cyl)) +  -->
<!--   scale_colour_brewer(palette  = 'Set1') + -->
<!--   scale_size (breaks = c(40,60))    -->

<!-- ggplot(mpg) +  -->
<!--   geom_point(aes(x = displ, y = hwy, colour =  class, size=cyl)) +  -->
<!--   scale_colour_brewer(palette  = 'Set1', breaks = c(4,6)) + -->
<!--   scale_size (breaks = c(4,6))    -->

<!-- ``` -->

- using `guides` to modify the `scale_` 
```{r}
 ggplot(mpg) +
   geom_point(aes(x = displ, y = hwy, colour = cyl, size=cyl)) +
   # scale_colour_brewer(palette  = 'Set1') +   #can not continuous 
   scale_size (breaks = c(4,5,6))  +
  
   guides( size = guide_legend(
            override.aes = list(color = c('red', 'blue', 'black'))))
```
 

- unite legends 

when multiple aesthetics are mapped to the same variable.


```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = cyl, size=cyl))   +
  guides(colour="legend")
```

`category is also ok` 
```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = fl, size=fl))   +
  guides(colour="legend") #size is not ok
```
 
 
 - x and y also have associated scales 

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  scale_x_continuous(breaks = c(3.5, 5, 6)) + 
  scale_y_continuous(trans = 'log2')
``` 
 
 


### Change coordinates

Changing the coordinate system can have dramatic effects 

- coord_polar

```{r}
ggplot(mpg) + 
  geom_bar(aes(x = class)) + 
  coord_polar()
```

```{r}
ggplot(mpg) + 
  geom_bar(aes(x = class)) + 
  coord_polar(theta = 'y') + 
  expand_limits(y = 70)
```

- specify the scale of coordinate 

```{r}
require(scales)
ggplot(mpg) + 
  geom_bar(aes(x = class)) + 
  scale_y_continuous(limits = c(0, 50), breaks = seq(0, 50, 01)) 
  
  # scale_y_continuous(labels = percent) # labels as percents
  # +
  # scale_x_discrete(labels=c(1:7)  )
```

- using coord_cartesian `zoom in` 
 
```{r}
# have been deleted 
ggplot(mpg) + 
  geom_bar(aes(x = class)) + 
  scale_y_continuous(limits = c(0, 30))+
  scale_x_discrete( limit=c("midsize","compact"))
```

```{r}
ggplot(mpg) + 
  geom_bar(aes(x = class)) + 
  coord_cartesian( ylim = c(0, 30))+
  scale_x_discrete( limit=c("midsize","compact"))
```
 

- reverse direction of axes

```{r}
ggplot(mpg) + 
  geom_point(aes(x = hwy, y = displ))+
  
  scale_x_continuous(breaks = c(20, 30, 35,40)) + 
  scale_y_reverse()+
  scale_y_continuous(trans="log2")
  
# log10, sqrt, reverse, scale_y_continuous(trans="log2")
```
 

### Customize axis ticks
- change axis text font, color, size, angle using theme 
```{r}
# when use theme, scale, guide? 
ggplot(mpg) + 
  geom_point(aes(x = hwy, y = displ))+
  
  theme(axis.text.x = element_text(face="bold", color="#993333", 
                           size=14, angle=45),
          axis.text.y = element_text(face="bold", color="blue", 
                           size=7, angle=90))
```

- `remove aixs ticks and tick labels`
```{r}
ggplot(mpg) + 
  geom_point(aes(x = hwy, y = displ))+
theme(
  axis.text.x = element_blank(), # Remove x axis tick labels
  axis.text.y = element_blank(), # Remove y axis tick labels
  axis.ticks = element_blank()) # Remove ticks
```

### Flip and reverse plot 


- boxplot and violin 
```{r}
ggplot(mpg ) + 
  geom_violin( aes(x = as.factor(cyl), y=hwy ,color=as.factor(cyl) ) ,trim = FALSE,width = 4 ) +
  geom_boxplot( aes(x = as.factor(cyl), y=hwy  ), notch = F , width = 0.1)  
 
```

- dotplot
`using stat_ function` 
```{r}
ggplot(mpg ,aes(x = as.factor(cyl), y=hwy ) )+ 
  geom_dotplot(aes(color =as.factor(cyl), fill = as.factor(cyl)), 
                 binaxis = "y", stackdir = "center") + 
  stat_summary(fun.data="mean_sdl" )  
```

- errorbar
```{r,echo=FALSE}
# Calculate the mean and the SD in each group
#+++++++++++++++++++++++++
# data : a data frame
# varname : the name of the variable to be summariezed
# grps : column names to be used as grouping variables
data_summary <- function(data, varname, grps){
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      sd = sd(x[[col]], na.rm=TRUE))
  }
  data_sum<-ddply(data, grps, .fun=summary_func, varname)
  data_sum <- rename(data_sum, c("mean" = varname))
 return(data_sum)
}

```

```{r}
df3 <- data_summary(mpg, varname="hwy", grps= c("cyl" ))
head(df3)
```


```{r}
ggplot( df3, aes(as.factor(cyl) ,  (hwy), ymin = hwy-sd, ymax = hwy+sd)  ) +
  geom_line(aes(group = 0 )) +
  geom_errorbar(aes(color = as.factor(cyl) ),width = 0.2) 
```
```{r}
ggplot( df3, aes(as.factor(cyl) ,  (hwy), ymin = hwy-sd, ymax = hwy+sd)  ) +
  geom_bar(aes(fill = as.factor(cyl)), stat = "identity",
             position = "dodge") +
  geom_errorbar(aes(   ),width = 0.2) +
  coord_flip()
```

*using original data* 
```{r}
 
  ggplot(mpg, aes(cyl, hwy)) +
  stat_summary(geom = "bar") +
  stat_summary(geom = "errorbar")
```


`flip`
```{r}
ggplot(mpg) + 
  geom_bar(aes(x = class)) +
  coord_flip()
```


`reverse`
```{r}
ggplot(mpg) + 
  geom_bar(aes(x = class)) +
  scale_y_reverse()
```


### Create stats

*Every geom has a stat. The stat can be overwritten if we use any additional computations.*

```{r}
ggplot(mpg ) + 
  geom_bar (aes(x = cyl ),position = 'identity' ) #using original data
```

- using transformed variables
```{r,message=FALSE,warning=FALSE}
library(dplyr)
library(ggplot2)
# mpg_counted <-    
#   count(mpg,  cyl  )
# head(mpg_counted)

# ggplot(mpg_counted) + 
#   geom_smooth(aes(x =  cyl , y = n)) +
#   geom_bar (aes(x =  cyl , y = n), stat = 'identity') #using summary data 
```

- using the `after_stat()` function inside `aes()`.  

```{r}
require(scales)
ggplot(mpg) + 
  geom_bar(aes(x = class, y = after_stat(  count / sum(count))))+
  scale_y_continuous(labels = percent) # labels decimals as percents
  
```

- using `density` geometric in histogram 

```{r}
ggplot(mpg,aes(x = hwy)) + 
  geom_histogram(aes(y=..density..))+
  geom_density(  ) 

```

 
- Use `stat_summary()` to add the mean of `hwy` for each group

`STAT vs. GEOM`
```{r}
p1 <- ggplot(mpg,aes(x = class, y = hwy) ) + 
  stat_summary(
    geom = "pointrange",
    fun.data = mean_se
  )

p2 <- ggplot(mpg,aes(x = class, y = hwy) ) + 
  stat_summary( 
  )

p3 <- ggplot(mpg,aes(x = class, y = hwy) ) + 
  stat_summary( 
  )+
  stat_summary(
    fun.data = ~mean_se(., mult = 1.96), # Increase `mult` value for bigger interval!
    geom = "errorbar",
  )

library(patchwork)
p1+p2+p3
```


```{r}
ggplot(mpg) + 
  geom_point(aes(x = class, y = hwy), width = 0.2)+
  stat_summary(aes( x = class,y = hwy), geom="point",color="red",size=4)
 
```

`jitter points`
```{r}
ggplot(mpg) + 
  geom_jitter(aes(x = class, y = hwy), width = 0.2)+
  stat_summary(aes( x = class,y = hwy), geom="point",color="red",size=4)
 
```


### Facets
`facet_wrap()` allows you to place facet side by side into a rectangular layout. `facet_grid()` allows you to specify different directions and works on two variables. 

- share the axes between the different panels 
```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~ class)
```

  
```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_grid(drv~ .  )
```

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_grid(~ drv  )
```
```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_grid(year ~ drv)
```

- do not share the axes between the different panels 
```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~ drv , scales = "free")
```

`only free y axes`
```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~ drv , scales = "free_y")
```
 
- adjust y scale (space) between the panels 

```{r}
ggplot(mpg) + 
  geom_bar(aes(y = manufacturer)) + 
  facet_grid(class ~ .)
```

```{r}
ggplot(mpg) + 
  geom_bar(aes(y = manufacturer)) + 
  facet_grid(class ~ ., space = "free_y", scales = "free_y")
```

- display by adding multiple variables together 

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(year ~ drv)
```

### Theme
 
`theme_minimal`
```{r}
ggplot(mpg) + 
  geom_bar(aes(y = class)) + 
  facet_wrap(~year) + 
  theme_minimal()
```

`Further adjustments`  
`theme_bw` 
```{r,message=FALSE,warning=FALSE}
ggplot(mpg) + 
  geom_bar(aes(y = class)) + 
  facet_wrap(~year) + 
  
  labs(title = "Number of car models per class",
       caption = "source: http://fueleconomy.gov",
       x = NULL,
       y = NULL) +
   
  theme_bw() + 
  theme(
 
    strip.text = element_text(face = 'bold', hjust = 0),
    
    plot.caption = element_text(face = 'italic'),
    
    panel.grid.major = element_line('white', size = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
    # ,
    # panel.ontop = TRUE
  )
```

`theme_classic`
```{r}
ggplot(mpg) + 
  geom_bar(aes(y = class, fill = drv)  ,position = "dodge") + 
  facet_wrap(~year) + 
  labs(title = "Number of car models per class",
       caption = "source: http://fueleconomy.gov",
       x = 'Number of cars',
       y = NULL)+
  
  scale_fill_manual(name = "Drive Models",values=c("black", "grey50",  "grey80"), labels = c("4w","Fw","Rw" )) +
  
  # scale_x_continuous(expand = c(0, NA)) + 
  
  theme_classic() + 
  theme(
    # text = element_text('Avenir Next Condensed'),
    # strip.text = element_text(face = 'bold', hjust = 0),
    
    plot.caption = element_text(face = 'italic'),
    
    panel.grid.major = element_line('white', size = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
    
    # panel.ontop = TRUE)
  )
```

- using ggthemes
```{r}
library(ggthemes)
ggplot(mpg) + 
  geom_bar(aes(y = class, fill = drv)  ,position = "dodge") + 
  facet_wrap(~year) + 
  labs(title = "Number of car models per class",
       caption = "source: http://fueleconomy.gov",
       x = 'Number of cars',
       y = NULL)+
  
  ggthemes::theme_economist()
```

- customized theme
```{r}
ggplot(mpg) + 
  geom_bar(aes(y = class, fill = drv)  ,position = "dodge") + 
  facet_wrap(~year) + 
  labs(title = "Number of car models per class",
       caption = "source: http://fueleconomy.gov",
       x = 'Number of cars',
       y = NULL)+
  
theme(
  
    # 1 change legend
    legend.background = element_rect(
      fill = "#fff6c2",
      color = "black",
      linetype = "dashed"
    ),
    legend.key = element_rect(fill = "grey", color = "brown"),
    legend.position = "bottom", 
    
    # 2 change panel (middle erea) background 
    panel.background = element_rect(
      fill = "#005F59",
      color = "red", size = 3
    ),
    panel.border = element_rect(
      color = "black",
      fill = "transparent",
      linetype = "dashed", size = 3
    ),
    
    # 3 change plot background
    plot.background = element_rect(
      fill = "#a1dce9",
      color = "black",
      size = 1.3
    ),
    
   
    
    # 4 change axis elements
    axis.line = element_line(color = "orange", size = 2),
    axis.title = element_text(color = "red", face = "italic"),
    axis.ticks = element_line(color = "purple", size = 3),
    axis.text = element_text(color = "blue"),
    axis.text.x = element_text(angle = 45, hjust = 1), 
    
    # 5 change facet panel
    strip.background = element_rect(fill = "orange"),
    strip.text = element_text(color = "red"),
    panel.spacing = unit(0.3, "inch") 
  )
```


### How to setup subscripts or superscripts

[see here](https://rpubs.com/Daniel_He/1041883)

## How to create advanced plots
[see here](https://rpubs.com/Daniel_He/941386)
