# Basic statistics

<!-- ============================= -->
## The essentials of R
### Manipulation of vector 

```{r,warning=FALSE}
library(tidyverse)
library(dplyr)
vec <- c(3,5,2,1,5,"O",NA)
length(unique(vec))

num_vec <- as.numeric(vec)
log(num_vec)

sum(c(num_vec, NA), na.rm=T)
sort(num_vec, decreasing = T)

is.na(num_vec)
num_vec[!is.na(num_vec)]

c(5,6) %in% vec
grepl("5", vec)
```
### Generate sequence or repeted sequece  

```{r}
seq(from = 0, to = 10, by = 0.5)
rep(x = 1:3, times = 4)
rep(x = 1:3, each = 4)
```
### Get directory and write data out and in

```{r}
getwd()
setwd(getwd())
write.csv(cars, "cars.csv", row.names=F)
dataframe  <- read.csv("cars.csv")

```
### Function

```{r}
my_func <- function(x){
  x_mod <- (x + 7) * 4
  return(x_mod)
}

my_func(num_vec)
```
### Plot 

```{r}
plot(dist ~ speed, data=cars)
hist(cars$dist )

```
### Build model and plot

```{r}
model <- lm(dist ~ speed, data=cars)
plot(dist ~ speed, data=cars)
abline(model)
abline(v = 25)
abline(h = 15)
```
### Rename names of columns 

```{r}
names(cars)
names(cars) <- c("speed per hour", "total dist")
```
### Class of dataframe 

```{r}
matrix <- as.matrix(cars)
df <- as.data.frame(matrix)
class(matrix)
class(df)

# tranform
t(matrix)
```
### Generate new variable for dataframe (character)

```{r}
paste0("raster_", 1:10)
paste0("raster_", rep(x = 1:5, times = 10))
df$group <- paste0("raster_", rep(x = 1:5, times = 10))
df$id <-  paste0("raster_",  1:50)
```
### Create a new dataframe using 'rnorm' - random  number from distribution

```{r}
sample <-  round((rnorm(50,0, 1)),2)
group <- paste0("raster_", rep(x = 1:5, times = 10))

df_join <- data.frame(sample, group)
df_join$id <-  paste0("raster_",  1:50)
```

### Left join two dataframes  
```{r}
library(dplyr)
data_all <- left_join(df, df_join, by="id")
head(data_all)
```
### Select variables
```{r}
select(data_all, group.x, id  )
```
### Filter observations
```{r}
raster_1 <- filter(data_all, group.x == "raster_1")
raster_1
speed_dist <- filter(data_all, data_all$`speed per hour` < 11 & data_all$`total dist` >= 10)
speed_dist
 
```
### Append rows
```{r}
rbind(raster_1,speed_dist)
```
 
### Create new variables instead of old variables
```{r}
mutate(data_all, 
       sample = round(sample,1))
```
### summarise statistics
```{r}
 summarise(data_all,
          mean_speed = mean(sample),
          max_dist = max( "total dist" ))
```
### Group dataframe then summarise statistics
```{r}
data_all_group <-   group_by(data_all, group.x)   
 summarise(data_all_group, 
          mean_speed = mean(sample),
          max_dist = max( "total dist" ))
```
### Ungroup then summarise statistics 
```{r}
ungroup_data <- ungroup( data_all_group)
 summarise(  ungroup_data , 
          mean_speed = mean(sample),
          max_dist = max( "total dist" ))

```
### Summary linear regression model
```{r}
mod1 <- lm(cars$`total dist` ~ cars$`speed per hour` )
summary(mod1) 
```
### Create frequency table
```{r}
table(data_all_group$`speed per hour`,data_all_group$group.x  )
```
### Value and variable label
```{r}
table(iris$Species)
```
```{r}
iris$Species <- factor(iris$Species,labels = c( "setosanew","versicolornew","virginianew"))
table(iris$Species)
```
```{r message=FALSE}
library(Hmisc)
```

```{r}
label(iris$Species) <- "Species types"
table(iris$Species)
```
### Recode a variable 
```{r}
irisifelse <-  iris%>% 
mutate(Sepal.Length2 = ifelse(Sepal.Length < 6 , "level1", ifelse(Sepal.Length < 7 , "level2", Sepal.Length)))

table(irisifelse$Sepal.Length2)
```