

## Linear mixed model covariance decomposition with random slopes- lme4

### Load data
```{r}
data(Orthodont, package="nlme")
Data <- Orthodont
```
  
### Using lmm

```{r}
library(lme4) 
fit.lmer.slope <- lmer(distance ~ age + Sex  +(1+ age | Subject), data = Data) 
summary(fit.lmer.slope)
```

**default is compound symmetry**

### Get x,y,z 
```{r}
X=(getME(fit.lmer.slope, "X"))
head(X)
y=getME(fit.lmer.slope, "y")
head(y)

Z <- getME(fit.lmer.slope, "Z")
head(Z)
dim(Z)
```

### Get fixed effect and random effect coefficients
```{r}
bhat <- getME(fit.lmer.slope, "fixef") #getME(fit.lmer, "beta")  # fixef(fit.lmer)
bhat 
uhat <- ranef(fit.lmer.slope) 
uhat
 
```

### Get random effect covariance structure (covariance matrix expandation)      

**`notice z matrix structure`**

<!-- # ```{r} -->
<!-- # vc <- VarCorr(fit.lmer.slope)  -->
<!-- # up <- vc[["Subject"]][1]*diag(ncol(Z))[ c(1:(ncol(Z)/2),0), ] -->
<!-- # down <- vc[["Subject"]][4]*diag(ncol(Z))[ c((ncol(Z)/2+1):(ncol(Z)),0), ] -->
<!-- # Lambda_new <- rbind(up,down) -->
<!-- #   -->
<!-- # head(Lambda_new) -->
<!-- # dim(Lambda_new) -->
<!-- # ``` -->

```{r}
vc <- VarCorr(fit.lmer.slope)
Lambda_new <- bdiag(replicate(length(levels(Data$Subject)), vc[["Subject"]], simplify = FALSE))
head(Lambda_new)
dim(Lambda_new)
```

- **"theta": random-effects parameter estimates: these are parameterized as the relative Cholesky factors of each random effect term**

- **use other cov structure by nlme**

- **lme4 does not allow to specify it, and it can only fit LMMs with independent residual errors. However, the range of available variance-covariance matrices for the random effects are restricted to diagonal or general matrices. **

<!--   Get  ST    -->
<!-- ```{r} -->
<!-- ST <- getME(fit.lmer, "ST") # $r$  -->

<!-- ``` -->
<!-- https://rdrr.io/cran/lme4/man/getME.html -->
<!-- https://en.wikipedia.org/wiki/Cholesky_decomposition -->

<!--   Get lower   -->
<!-- ```{r} -->
<!-- lower <- getME(fit.lmer, "lower") # $r$  -->
<!-- lower -->
<!-- ``` -->

### Get residual variance and its identity matrix
```{r}
sigma <- getME(fit.lmer.slope, "sigma")**2 
sigma
head(sigma*diag(nrow(Data)))
 
```

### Get y covariance matrix  

```{r}
VM <- Z%*%Lambda_new%*%t(Z)+sigma*diag(nrow(Data)) 
head(VM)
dim(VM)
```

### Get fixed effect coefficients covariance matrix
```{r}
vcov <- vcov(fit.lmer.slope) #fixed cov
vcov
# computer correlation coefficients 
vcov@x[2]/prod(sqrt( diag(vcov(fit.lmer.slope))[-3] ))
```

### Get random effect coefficients covariance matrix (standard deviation)
```{r}
vc <- VarCorr(fit.lmer.slope) 
vc ## default print method: standard dev and corr
as.matrix(Matrix::bdiag(vc)) #random effect covariance matrix
 
# https://stackoverflow.com/questions/47307340/extracting-the-i-estimated-variance-covariance-matrix-of-random-effects-and-or
# https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mixed_sect022.htm
```


### Computer fixed effect coefficients

```{r}
library(matlib)
inv(t(as.matrix(X))%*%inv(as.matrix(VM))%*%as.matrix(X))%*%t(as.matrix(X))%*%inv(as.matrix(VM))%*%y
bhat
```

### Computer covariance of fixed efffect coefficients
```{r}
inv(t(as.matrix(X))%*%inv(as.matrix(VM))%*%as.matrix(X))
# standard error
sqrt(inv(t(as.matrix(X))%*%inv(as.matrix(VM))%*%as.matrix(X)))

# the following equals lmm summary
vcov(fit.lmer.slope) 
sqrt(vcov(fit.lmer.slope))
```
**default is compound symmetry**

### Computer random effect coefficients 
```{r}
comput_uhat <- (as.matrix(Lambda_new))%*%t(Z)%*%inv(as.matrix(VM))%*%(y-as.matrix(X)%*%(bhat))
cbind((comput_uhat@x),(uhat[["Subject"]]))

```

- covariance matrix of random effect coefficients
```{r}
head(Lambda_new)
```
 

### Computer predicted values
```{r}
yhat <- X%*%(bhat)+Z%*%comput_uhat
# comput_uhat= uhat
head(yhat)
head(fitted(fit.lmer.slope))
```


### Using lmm with two grouping factors

```{r}
library(lme4) 
fit.lmer.slope2 <- lmer(distance ~ age + Sex  +(1+ age | Subject)+ (1 | age), data = Data) 
summary(fit.lmer.slope2)
```
 
### Get  z2 

`58= 54+4 ages `
```{r}
Z2 <- getME(fit.lmer.slope2, "Z")
head(Z2,10)
dim(Z2)
```

### Using lmm with nested or crossed random effects 

```{r}
library(lme4) 
fit.lmer.slope3 <- lmer(distance ~ age + Sex  +(1 | Sex/Subject), data = Data) 
summary(fit.lmer.slope3)
```
 
### Get  z3 

`29= 27+2 ages `
```{r}
Z3 <- getME(fit.lmer.slope3, "Z")
head(Z3,10)
dim(Z3)
```

`Remark`

nested random effects occur when one factor (grouping variable) appears only within a particular level of another factor (grouping variable).  
lmer 
(1|group1/group2)
 
Crossed random effects are simply: not nested. individual observations are nested separately within the two or more factors.  
lmer 
(1|group1) + (1|group2)
