## Linear Mixed Model Covariance Decomposition with Random Slopes (lme4)

In this section, we extend the covariance decomposition of linear mixed models (LMMs) from **random intercept models** to **random slope models**. Allowing random slopes means that, in addition to subject-specific baselines (intercepts), the effect of a covariate (here, age) is also allowed to vary across subjects. This is particularly important in longitudinal and growth-curve settings, where individuals may follow systematically different trajectories over time.

We continue to use the **Orthodont** dataset, a classic longitudinal dataset that records the distance of orthodontic measurements over age, stratified by subject and sex.

---

### Load data

We begin by loading the Orthodont dataset from the `nlme` package. Each subject has repeated measurements over age, making this dataset well suited for mixed-effects modeling.

```{r}
data(Orthodont, package="nlme")
Data <- Orthodont
```

---

### Using a linear mixed model with random slopes

We now fit a linear mixed model with:
- fixed effects for **age** and **Sex**,
- a **random intercept** for each Subject,
- and a **random slope for age** within each Subject.

Mathematically, this model can be written as

$$
y_{ij} = (\beta_0 + u_{0j}) + (\beta_1 + u_{1j}) \, \text{age}_{ij} + \beta_2 \, \text{Sex}_{ij} + \varepsilon_{ij},
$$

where:
- $(u_{0j}, u_{1j})$ are subject-specific random effects,
- $\varepsilon_{ij} \sim N(0, \sigma^2)$ are residual errors.

```{r}
library(lme4) 
fit.lmer.slope <- lmer(distance ~ age + Sex  +(1+ age | Subject), data = Data) 
summary(fit.lmer.slope)
```

Although the residual covariance is independent (diagonal), the **random-effects covariance structure** is no longer scalar. Instead, each subject has a **2×2 covariance matrix** corresponding to the random intercept and random slope.

---

### Extracting the design matrices X, y, and Z

To understand the internal structure of the fitted model, we explicitly extract:
- **X**: the fixed-effects design matrix,
- **y**: the response vector,
- **Z**: the random-effects design matrix.

The Z matrix is particularly important here because it now contains **two columns per subject**: one for the random intercept and one for the random slope.

```{r}
X=(getME(fit.lmer.slope, "X"))
head(X)
y=getME(fit.lmer.slope, "y")
head(y)

Z <- getME(fit.lmer.slope, "Z")
head(Z)
dim(Z)
```

Each block of columns in Z corresponds to one subject, and within each block, the columns represent the intercept and slope contributions.

---

### Fixed and random effect coefficients

We next extract:
- $\hat{\beta}$, the estimated fixed-effect coefficients,
- $\hat{u}$, the estimated random effects (BLUPs).

The random effects are returned as a list indexed by grouping factor.

```{r}
bhat <- getME(fit.lmer.slope, "fixef") # fixed effects
bhat 
uhat <- ranef(fit.lmer.slope) 
uhat
 
```

Each subject now has **two random-effect coefficients**: one intercept deviation and one slope deviation.

---

### Random-effect covariance structure (matrix expansion)

In a random slope model, the random-effects covariance matrix for a single subject is:

$$
\mathbf{G}_{\text{subject}} =
\begin{pmatrix}
\sigma^2_{\text{int}} & \sigma_{\text{int,slope}} \\
\sigma_{\text{int,slope}} & \sigma^2_{\text{slope}}
\end{pmatrix}.
$$

The full covariance matrix $\mathbf{G}$ is block-diagonal, with one such matrix per subject. We explicitly construct this block-diagonal matrix using `bdiag()`.

```{r}
vc <- VarCorr(fit.lmer.slope)
Lambda_new <- bdiag(replicate(length(levels(Data$Subject)), vc[["Subject"]], simplify = FALSE))
head(Lambda_new)
dim(Lambda_new)
```

Here:
- `VarCorr()` provides the estimated standard deviations and correlations,
- `bdiag()` assembles the subject-specific covariance matrices into a global random-effects covariance matrix.

> **Note**  
> In `lme4`, the internal parameters `theta` correspond to the Cholesky factors of these covariance matrices. Given `theta`, the full covariance matrix is reconstructed implicitly.

---

### Residual variance and identity matrix

The residual errors are assumed independent with variance $\sigma^2$. We extract $\sigma^2$ and explicitly form $\sigma^2 I$.

```{r}
sigma <- getME(fit.lmer.slope, "sigma")**2 
sigma
head(sigma*diag(nrow(Data)))
 
```

---

### Covariance matrix of the response vector y

The marginal covariance of $\mathbf{y}$ is:

$$
\mathbf{V} = \mathbf{Z G Z}^\top + \sigma^2 \mathbf{I}.
$$

We compute this matrix explicitly.

```{r}
VM <- Z%*%Lambda_new%*%t(Z)+sigma*diag(nrow(Data)) 
head(VM)
dim(VM)
```

This matrix fully characterizes the dependence structure induced by the random intercepts and random slopes.

---

### Covariance matrix of fixed-effect coefficients

The covariance matrix of $\hat{\beta}$ is given by:

$$
\operatorname{Var}(\hat{\beta}) = (\mathbf{X}^\top \mathbf{V}^{-1} \mathbf{X})^{-1}.
$$

We confirm that this matches the output from `vcov()`.

```{r}
vcov <- vcov(fit.lmer.slope) # fixed-effect covariance
vcov
# compute correlation coefficient
vcov@x[2]/prod(sqrt( diag(vcov(fit.lmer.slope))[-3] ))
```

---

### Random-effect covariance matrix

The block-diagonal random-effects covariance matrix can also be extracted directly.

```{r}
vc <- VarCorr(fit.lmer.slope) 
vc ## standard deviations and correlations
as.matrix(Matrix::bdiag(vc)) # random-effect covariance matrix
```

---

### Computing fixed-effect coefficients manually

Using the generalized least squares (GLS) formula,

$$
\hat{\beta} = (\mathbf{X}^\top \mathbf{V}^{-1} \mathbf{X})^{-1}\mathbf{X}^\top \mathbf{V}^{-1}\mathbf{y},
$$

we compute $\hat{\beta}$ explicitly and verify it matches the model output.

```{r}
library(matlib)
inv(t(as.matrix(X))%*%inv(as.matrix(VM))%*%as.matrix(X))%*%t(as.matrix(X))%*%inv(as.matrix(VM))%*%y
bhat
```

---

### Covariance of fixed-effect coefficients

We also compute the covariance matrix and standard errors manually.

```{r}
inv(t(as.matrix(X))%*%inv(as.matrix(VM))%*%as.matrix(X))
# standard errors
sqrt(inv(t(as.matrix(X))%*%inv(as.matrix(VM))%*%as.matrix(X)))

# matches lmer output
vcov(fit.lmer.slope) 
sqrt(vcov(fit.lmer.slope))
```

---

### Computing random-effect coefficients

The BLUPs of the random effects are given by:

$$
\hat{\mathbf{u}} = \mathbf{GZ}^\top \mathbf{V}^{-1} (\mathbf{y} - \mathbf{X}\hat{\beta}).
$$

We compute these manually and compare them with `ranef()`.

```{r}
comput_uhat <- (as.matrix(Lambda_new))%*%t(Z)%*%inv(as.matrix(VM))%*%(y-as.matrix(X)%*%(bhat))
cbind((comput_uhat@x),(uhat[["Subject"]]))
```

---

### Predicted values

Predicted values are obtained by combining fixed and random components:

$$
\hat{\mathbf{y}} = \mathbf{X}\hat{\beta} + \mathbf{Z}\hat{\mathbf{u}}.
$$

```{r}
yhat <- X%*%(bhat)+Z%*%comput_uhat
head(yhat)
head(fitted(fit.lmer.slope))
```

---

### Model with two grouping factors

We now illustrate a model with **two random-effect grouping factors**, adding a random intercept for `age` as well.

```{r}
library(lme4) 
fit.lmer.slope2 <- lmer(distance ~ age + Sex  +(1+ age | Subject)+ (1 | age), data = Data) 
summary(fit.lmer.slope2)
```

---

### Extracting Z for the two-factor model

Here, the Z matrix expands accordingly. The dimension reflects the combined random effects.

```{r}
Z2 <- getME(fit.lmer.slope2, "Z")
head(Z2,10)
dim(Z2)
```

---

### Nested versus crossed random effects

Finally, we demonstrate a nested random-effects structure, where subjects are nested within sex.

```{r}
library(lme4) 
fit.lmer.slope3 <- lmer(distance ~ age + Sex  +(1 | Sex/Subject), data = Data) 
summary(fit.lmer.slope3)
```

The corresponding Z matrix reflects this nesting.

```{r}
Z3 <- getME(fit.lmer.slope3, "Z")
head(Z3,10)
dim(Z3)
```

---

### Remark: Nested vs. crossed random effects

- **Nested random effects** occur when one grouping factor appears only within levels of another:

(1 | group1/group2)


- **Crossed random effects** occur when observations are independently grouped by multiple factors:


(1 | group1) + (1 | group2)


Understanding how these structures affect the Z matrix and covariance decomposition is essential for correctly specifying and interpreting linear mixed models in practice.
