
# SASmarkdown

## How to install sasmarkdown 
[sasmarkdown](https://www.ssc.wisc.edu/~hemken/SASworkshops/Markdown/SASmdpackage.html) 

 `Use an Engine`
```{r libload} 
# knitr::opts_chunk$set(echo = TRUE)
require(SASmarkdown)
```

```{r}
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -ls 75"
knitr::opts_chunk$set(engine='sashtml', engine.path=saspath,
                      engine.opts=sasopts, comment="")
```

## Common statements 
`Generally, can not use r syntax in these chunks ` 

`Data can not be used in different chunks`

### read in (date) data using informat
```{r , engine="sas" , error=TRUE}
/*import and export raw data using infile*/
data test;

/*infile "c:/document/data.text";*/
input name $ height weight;
length height 4;
cards 
;
Daniel 173 150
;
run;
proc print;run;
```

```{r , engine="sas" , error=TRUE}
data have;
input char_date : $20. fmt : $20.;
/* set a new variable format*/
num_char=inputn(char_date,fmt);
format num_char date9. ;
cards;
12-08-2015  DDMMYY12.
12/8/2016   MMDDYY12.
05/25/2015  MMDDYY12.
;
run;
proc print;run;
```
```{r , engine="sas" , error=TRUE}
data have;
input char_date MMDDYY10.  ;
/* output format */
format char_date date9. ;
cards;
12-08-2015   
12/8/2016   
05/25/2015   
;
run;
proc print;run;
```

```{r , engine="sas" , error=TRUE}
data have;
/* set input and output format*/
input char_date MMDDYY10.  ;
format char_date MMDDYY10. ;
cards;
12-08-2015   
12/8/2016   
05/25/2015   
;
run;
proc print;run;
````

### compute mean and freqency 
```{r , engine="sas" , error=TRUE}
proc means data=sashelp.class maxdec=5 ;
run;

proc freq data=sashelp.class  ;
table sex;
run;

proc contents data=sashelp.class varnum  ;
run;

```

### sort a data set
```{r , engine="sas" , error=TRUE}
proc sort data=sashelp.class Out= name ;
by name;
Run;

proc print data= name (obs=10) ;
run;
```

### transpose or reshape

`wide format data to long format data`
```{r , engine="sas" , error=TRUE}
proc transpose data= sashelp.class out= Field;
by name;
Run;

proc print data= field (obs=10) ;
run;
```

`long format data to wide format data`
```{r , engine="sas" , error=TRUE}
proc transpose data= sashelp.class out= Field;
by name;
Run;

proc transpose data=Field out=Field_wide  ;
    by name ;
    id _name_;
    var col1;
run;

proc print data= field_wide (obs=10) ;
run;
```

### conditional statement
```{r , engine="sas" , error=TRUE}
 
DATA ab;
set sashelp.class;
IF sex="F" then message='A is greater';
Else message='B is greater';
Run; 

proc print data=ab (obs=10);
run;
```
```{r , engine="sas" , error=TRUE}
/*if else syntax*/
Data class;
set sashelp.class;
if weight >=100 then weight_cat =3;
else if weight >=90 and weight <100 then weight_cat =2 ;
else weight_cat =1;
attrib weight_cat label = "weight category,  1=Inad, 2=Adeq, 3=Exces"; 
run;

proc print data=class (obs=3);
run;
```
```{r , engine="sas" , error=TRUE}
/*if index*/
data test;
set sashelp.class;
length gender $10 ;
if index(sex,"M")>0 THEN gender="male";
if index(sex,"F")>0 THEN gender="female";
RUN;

proc print data=test (obs=5);
run;
```

### using like operation to select rows contaning a pattern
```{r , engine="sas" , error=TRUE}
/*using like operation to select rows contaning a pattern-matching text.*/
proc print data =sashelp.class;
where upcase(name) like '%A%';
run;

proc print data =sashelp.class;
where  (name) like '%A%';
run;
```

### change format of the variable 
```{r , engine="sas" , error=TRUE}
proc format;
   value AGEnew
        11 = '1: NEW' 
        12 = '2: NEW'
        13 = '3: NEW' 
        14 = '4: NEW'
        15 = '5: NEW'
        16 = '6: NEW'        
        ;
        run;
        
DATA ab;
set sashelp.class;
Format AGE AGEnew.;
Run;
        
proc freq data=ab;
        table AGE;
        run;
        
```
### basic operations
```{r  , engine="sas" , error=TRUE}
Data Mathdata;
A= 10.12345;
B=20;
C= mean (a,b);
D= Min(a,b);
E= Max(a,b); 
F = log(a);
G= round(a,0.02);
H= floor(a );

Run;

proc print data=Mathdata;
run;

Data mathdata;
Set  sashelp.Iris;
Sum = sum (of SepalLength, SepalWIDTH);
Diff = SepalLength- SepalWIDTH;
Mult = SepalLength* SepalWIDTH;
Div= SepalLength/ SepalWIDTH;
Run;

proc print data=mathdata (obs=10);
run;
```
### rename variables  

`a variable`
```{r  , engine="sas" , error=TRUE}
 Data AB;
set sashelp.class; Rename AGE= AGENEW;
Run;

proc print data=AB;
run;
```

`multiple variables`
```{r  , engine="sas" , error=TRUE}
/*rename*/
data AAA;
 set sashelp.class;
run;

/*create a list*/
proc sql noprint;
   select cats(name,'=',name,'_new')
          into : lis
          separated by ' '
          from dictionary.columns
          where libname = 'WORK' and memname = 'AAA';
quit;
%put &lis;

proc datasets library = work nolist;
   modify AAA;
   rename &lis;
quit;

proc print data=aaa (obs=2);
run;    
          
/*function to deal with text*/
%scan(&lis,2);   
/*#scan variable*/
%substr(NAME, %length(NAME),1);  
/*#extract character*/
%substr(NAME, 2,1);    
%index(&lis,"Age=Age_new") ;         
/*#identify this exists  */
```

### extract text from a character  value
```{r  , engine="sas" , error=TRUE}
Data Mathdata;
Text = "Hello World";
Text1= substr(Text, 6, 2);
Run;

proc print data=mathdata  ;
run;
```
### convert the character value into a numeric value and reverse
```{r , engine="sas" , error=TRUE}
Data ABC;
set sashelp.class;
agenew= input (age, best.); 
Run;

proc print data=abc  ;
run;

```

```{r , engine="sas" , error=TRUE}
Data ABC;
set sashelp.class;
agenew= put (age, best.); 
Run;

proc contents data=abc  ;
run;
```

### change the length of the variable
```{r , engine="sas" , error=TRUE}
Data ABC;
set sashelp.class;
Length agenew $10.;
Label agenew=“New age”;
Run;

proc contents data=abc  ;
run;
 
```
### create a report
```{r , engine="sas" , error=TRUE}
proc report data=sashelp.class;
Column age;
Define age / display;
Run;
```

### create a random variable
```{r , engine="sas" , error=TRUE}
data ab;
set sashelp.class;
num=rand("normal");
run;

proc print data=ab (obs=10); 
run;
```
### combine two texts
```{r , engine="sas" , error=TRUE}
Data Mathdata;
Text = "Hello";
Text1= "World";
Text2= text || " " ||text1;
Run;

proc print data=Mathdata  ; 
run;
```
### compress spaces
```{r , engine="sas" , error=TRUE}
Data Mathdata;
Text = "Hello  World  ";
Text1= trim(text);
Text2= compress(text);
Run;

proc print data=Mathdata  ; 
run;
```

### identify the position of a specified text
```{r , engine="sas" , error=TRUE}
Data Mathdata;
Text = "Hello World";
indextext=  index(text, "or");
Run;

proc print data=Mathdata  ; 
run;
```
### convert upcase, lower case and propcase
```{r , engine="sas" , error=TRUE}
Data Mathdata;
Text = "Hello World";
upcase=  upcase(text );
lowcase=  lowcase(text );
propcase=  propcase(text );
Run;

proc print data=Mathdata  ; 
run;
```
### deduplication
```{r, engine="sas" , error=TRUE}
/* Dedup, original data has 19 observations  */
proc sort data = sashelp.class out =  dedup
nodupkeys;
by height; 
run;

proc print data= dedup;
run;

```
   
`The second way`
```{r, engine="sas" , error=TRUE}
proc sort data=sashelp.class out=data_rank;  by height  ; run;

data outdata dropdata;
 set data_rank;
 by height ;
 if last.height   then output  outdata;
else output dropdata; 
run;
proc print data=dropdata;run; 
```

### select sub data set
```{r,engine="sas", error=TRUE}
data where;
set sashelp.class;
where sex ne "F";
/*if */
run;

proc print data= where;
run;
```


### create macro using do loop statement 
```{r,engine="sas",error=TRUE}
/*create a macro to calcualte descriptive stats */
%macro means(var_avg) ;

/*calculate means*/
proc means data=sashelp.class StackODSOutput n mean std  min p5 p95 max nmiss; 
var  &var_avg;
class sex;
ods output summary=result2;
run;

/*append then output*/
data masterresult2;                  * combine results;
set masterresult2 result2;
run;

%mend means;

/*use macro to merge all descriptive stats */
data masterresult2 ;
set _null_;
run;

%let vars= 
age
height
weight
;

%macro model ;
%do i=1 %to %sysfunc(countw(&vars));

%let x=%scan(&vars,&i);
 %means( &x )
 
%end;
%mend model;

%model;

proc print data= masterresult2;
run;

```
### output intermediate tables
```{r,engine="sas",error=TRUE}
/*extract tables using ods trace*/
ods trace on;
proc freq data=sashelp.class  ;
table  sex;
run;
ods trace off;
```

```{r,engine="sas",error=TRUE}
ods   listing close;
ods trace on;
ods output ParameterEstimates= ParameterEstimates  ;
proc glm data=sashelp.class;
model height=age;
run;
ods trace off;
ods listing;

proc print data=ParameterEstimates;
run;

```

### creat a sequence number

`creat a sequence number based on a variable by grouping`

```{r,engine="sas",error=TRUE}

/* grouped*/
proc sort data=sashelp.class out=class ; by name; run; 
data temp; set class; 
	by name; 
  	count + 1;
	if first.name then count = 1; 
run; 
proc print data=temp (obs=3);run; 


proc sort data=class out=class2 ; by height; run; 
data class3;
set class2;
by height;
retain count;
if first.height then count=1;
else count=count+1;
run;
proc print data=class3 (obs=3);run; 

 /*nogrouped*/
data new;
 set class;
 seqno = _n_;
run;

proc print data=new (obs=3);run; 
```

### merge data sets

`right, left join`
```{r,engine="sas",error=TRUE}
data class2;
input name $   score;
datalines;
Alfred 	85
Alice 89
Daniel 99
;
run;

proc sql;
create table class_c as
  select *
    from sashelp.class as a
	  left join class2 as b 
	    on a.name = b.name;
quit;

proc print data=class_c;
run;
	    
proc sql;
create table class_d as
  select *
    from class2 as a
	  right join sashelp.class as b 
	    on a.name = b.name;
quit;

proc print data=class_d;
run;

```
`full join`
```{r,engine="sas",error=TRUE}
data class2;
input name $   score;
datalines;
Alfred 	85
Alice 89
Daniel 99
;
run;
 
proc sql;
create table class_e as
  select *
    from sashelp.class as a
	  full   join class2 as b 
	    on a.name = b.name;
quit;

proc print data=class_e;
run;

```


`or using merge for appending`
```{r,engine="sas",error=TRUE}
data class2;
input name $   score;
datalines;
Alfred 	85
Alice 89
Daniel 99
;
run;
 
data class_f;
merge  sashelp.class class2;
by name;
/*if A and B;  */
run;

proc print data=class_f;
run;

```

`inner join`
```{r,engine="sas",error=TRUE}
data class2;
input name $   score;
datalines;
Alfred 	85
Alice 89
Daniel 99
;
run;
 
proc sql;
create table class_e as
  select *
    from sashelp.class as a
	  inner join class2 as b 
	    on a.name = b.name;
quit;

proc print data=class_e;
run;

```

`minus join`
```{r,engine="sas",error=TRUE}
data class2;
input name $   score;
datalines;
Alfred 	85
Alice 89
Daniel 99
;
run;
 
proc sql;
create table class_e as
  select *
    from sashelp.class as a
	  left join class2 as b 
	    on a.name = b.name 
	    where b.name is NULL;
quit;

proc print data=class_e;
run;

/*proc sql ;*/
/*create table  arm as*/
/*select * from arm*/
/*where subject_id not in (select subject_id from ustwin);*/
/*quit;*/	    
```




### create table 1
```{r , engine="sas" , error=TRUE}
  *** Load utility macros;
%include "C:\Users\hed2\Downloads\mybook2\mybook2\sasmacro\create_table1.sas";

  *** Specify input and output data sets, and the column variable.;
%let INPUT_DATA  = sashelp.class;
%let OUTPUT_DATA = Table1 ;
%let COLVAR      = sex;

  /*chort*/
 *** %AddText(text=Height);
 *** %CategoricalRowVar2(rowVar=);
%ContinuousRowVar2(rowVar=height );
 *** %AddText(text=);

%ContinuousRowVar2(rowVar=weight );
%ContinuousRowVar2(rowVar=Age );

proc print data= table1;
run;

/* Export Table 1 as a CSV file*/ 
proc export data=Table1 replace label
    outfile="C:\Users\hed2\Downloads\mybook2\mybook2\sasmacro\table1.csv"
    dbms=csv;
run;  

```
### create table 1 without statistical test
```{r , engine="sas" , error=TRUE}
 *** Load utility macros;
%include "C:\Users\hed2\Downloads\mybook2\mybook2\sasmacro/Utility Macros_updated_NG.sas";

/* 
  By  counts 
%let yourdata = sashelp.class;
%let output_data= By_sex;
%let formatsfolder= ; 
%let yourfolder = ;
%let varlist_cat =  ;
%let varlist_cont = age heigt weight; 
%let output_order = height weight age;
%let decimal_max =3;
%let group_by = sex;
%let group_by_missing=0;

%Table_summary;
*/
```


### perform a regression 
```{r , engine="sas" , error=TRUE}
proc glm data= sashelp.class;
model height=age;
Run;
```


## Using macro
### Create a macro variable
```{r , engine="sas" , error=TRUE}

/*print mean of a variable to log in SAS*/
proc sql noprint; 
select  std(age) format=best32.
   into :age_mean
   from sashelp.class;
quit;
   
%put Mean of age: &age_mean;
%putlog Mean of age: &age_mean;
```

### Select variables names
```{r , engine="sas" , error=TRUE}

proc sql;
     select name
           into :vars separated by ' '
            from dictionary.columns
              where libname="SASHELP" and
                memname="CLASS" and varnum=4;
   ;
quit;
%put &vars.;
```

### Use iterative DO TO syntax to iterate a Specific Number  
```{r , engine="sas" , error=TRUE}
data do_to;
   x=10;
   y="yes";
   do i=1 to 10;
      x=x+1;
      output;
   end;
run;
proc print data=do_to;
run;
```

`use while or until`
```{r , engine="sas" , error=TRUE}

data loan;
   balance=1000;
   payment=0;
   do while (balance>0);
      balance=balance-100;
      payment=payment+1;
      output;
   end;
run;
proc print data=loan;
run;
```

`do loop within a macro`
```{r , engine="sas" , error=TRUE}

%macro run_calculation(amt, t, r);
    data customer_value;
        i=&r./10.;
        do n=0 to &t.;
            S=&amt.*((1+i)*n - 1)/i;  /*Power operator */
            output; /*output s*/
        end;
   file print;
   putlog  s @@;
   put  s @@;
    run;
proc print data=customer_value;
run;
%mend;

%run_calculation(amt=100, t=10, r=7);
```

### PROC REPORT allows more flexibility in displaying the data 
```{r , engine="sas" , error=TRUE}

proc means data=sashelp.class;
var _NUMERIC_;  /*_CHARACTER_*/
output out=want mean= sum= max= /autoname;
run;

proc report data= want; /*not use output but dataset*/
Column _FREQ_;
Column Age_Mean;
Column Age_Sum;
Column Age_max;
Define _FREQ_  /"The total number" display;
Define Age_mean  /"Mean of age" display;
Define Age_sum  /"Sum of age" display;
Define Age_max  /"Max of age" display;
Run;
```

### Create a macro to calculate descriptive stats and summarize tabels 
```{r , engine="sas" , error=TRUE}

%macro means(var_avg) ;

/*calculate means*/
proc means data=sashelp.class StackODSOutput n mean std  min p5 p95 max nmiss; 
var  &var_avg;
class sex;
ods output summary=result2;
run;

/*append then output*/
data masterresult2;                  * combine results;
set masterresult2 result2;
run;

%mend means;

/*use macro to merge all descriptive stats */
data masterresult2 ;
set _null_;
run;

%let vars= 
age
height
weight
;

%macro model ;
%do i=1 %to %sysfunc(countw(&vars));

%let x=%scan(&vars,&i);
 %means( &x )
 
%end;
%mend model;

%model;

proc print data= masterresult2;
run;
```

<!--  /*Tableau */ -->
<!-- /*https://sascrunch.com/members-area/*/ -->



## Use SAS format

### Creating labels for different values 
```{r , engine="sas" , error=TRUE}

PROC FORMAT;
    VALUE LIKERT7_A
	  1,2,3   = "Disagree"
	  4       = "Neither Agree nor Disagree"
	  5,6,7   = "Agree"
RUN;

PROC FORMAT;
    VALUE INCOME
	  LOW   -< 20000 = "Low"
	  20000 -< 60000 = "Middle"
	  60000 - HIGH   = "High";
RUN;

PROC FORMAT;
    VALUE RACE
       1     = "White"
       2     = "Black"
       OTHER = "Other";
RUN;

/*CREATING LABELS FOR CHARACTER VARIABLES*/
/*assigning labels to variable values*/
PROC FORMAT;
    VALUE $GENDERLABEL
	  "M"   = "Male"
	  "F"   = "Female";
RUN;


/*Use defined format*/
DATA sample;
    SET sashelp.class;
    FORMAT sex GENDERLABEL. ;
RUN;

proc freq data=sample;
table sex;
run;

```

### Storing farmats to a library- way 1
```{r , engine="sas" , error=TRUE}

LIBNAME format  "C:\Users\hed2\Downloads\code-storage\code\format";

PROC FORMAT LIBRARY=format ;
    VALUE $GENDER
	  "M"   = "Male_new"
	  "F"   = "Female_new";
RUN;
 
/*Use stored defined format way 1 */
OPTIONS FMTSEARCH=(format);
DATA sample;
    SET sashelp.class;
    FORMAT sex GENDER. ;
RUN;

proc freq data=sample;
table sex;
run;


```

### Use stored defined format- way 2 

```{r , engine="sas" , error=TRUE}
 

%INCLUDE 'C:\Users\hed2\Downloads\code-storage\code\format\format_test.sas';
DATA sample;
    SET sashelp.class;
    FORMAT sex GENDERsex. ;
RUN;

proc freq data=sample;
table sex;
run;

```

### Use stored defined format- way 3

`Add formats to existing formatlib; seperate creating labels, useing format and labeling variables.`
```{r , engine="sas" , error=TRUE}
 LIBNAME format  "C:\Users\hed2\Downloads\code-storage\code\format";

PROC FORMAT LIBRARY=format ;
    VALUE $GENDER
	  "M"   = "Male_new"
	  "F"   = "Female_new";
RUN;

	  
/*add formats to existing formatlib*/
PROC FORMAT LIBRARY=format ;
  VALUE $gendersextwo
	  "M"   = "Male_new2_two"
	  "F"   = "Female_new2_two";
RUN;
	  
/*use stored defined format way 3 */
OPTIONS FMTSEARCH=(format);
DATA sample;
    SET sashelp.class;
    FORMAT sex GENDERsextwo. ;
 RUN;

proc freq data=sample;
table sex;
run;

```

### Use stored defined format in batch- way 4 
```{r , engine="sas" , error=TRUE}

LIBNAME format  "C:\Users\hed2\Downloads\code-storage\code\format";

PROC FORMAT LIBRARY=format ;
    VALUE $GENDER
	  "M"   = "Male_new"
	  "F"   = "Female_new";
RUN;


/*use sas file to add format in batch*/
OPTIONS FMTSEARCH=(format);
DATA sample;
    SET sashelp.class;
%INCLUDE 'C:\Users\hed2\Downloads\code-storage\code\format\use_format.sas';
RUN;

proc freq data=sample;
table sex;
run;
```

### Modify format of the existing proc format 
```{r, engine="sas" , error=TRUE}
/*User-defined formats cannot be edited. to create a macro by myself ; or proc freq then use excel*/; 
PROC FORMAT ;
    VALUE $GENDER
	  "M"   = "Male_new"
	  "F"   = "Female_new";
RUN;
proc format;
	  value $sex_f  "F"="Female_f" other=[5.1];	
run;

DATA sample;
    SET sashelp.class;
    FORMAT sex $sex_f. ;
 RUN;

proc freq data=sample;
table sex;
run;
```

### Transform missing to char then to numeric again
```{r, engine="sas" , error=TRUE}
PROC FORMAT ;
    VALUE $num
	  "M"   = 1
	  "F"   = 2;
RUN;

DATA sample_2; 
  SET sashelp.class;
  sex_2 = input(put(sex ,num.),best.);  /*Converting variable types and keep*/
RUN;
proc freq data=sample_2;
table sex_2;
run;
```

### Output format of the existing format name by alphabet order 
```{r, engine="sas" , error=TRUE}

/*but we dont know the format name*/

/*PROC FORMAT FMTLIB;*/
PROC FORMAT ;
    VALUE $GENDER
	  "M"   = "Male_new"
	  "F"   = "Female_new";
RUN;

proc format library = work.formats FMTLIB cntlout = cntlout;
select $GENDER;
run;
```

### A macro to copy the existing proc format
```{r, engine="sas" , error=TRUE}
%macro formatd(data, var);
data temp;
set &data;
keep &var;
run;

/*original freq*/
ods output  OneWayFreqs=sample_b  ;
proc freq data=temp  ;
table  &var /missing ;
run;
ods output close;
 
/*delete format freq*/
data sample2;
set  temp;
Format _all_;
run;
proc freq data=sample2   ;
table  &var /missing out=sample_c ;
run;

/*select original variable code*/
proc sql noprint;
          select name into :firstvar_b from dictionary.columns where libname='WORK' and memname='SAMPLE_B'
          and varnum=2;
quit;
data sample_b2;
set sample_b;
Keep &firstvar_b ;
run;

/*select original variable label*/
proc sql noprint;
          select name into :firstvar_c from dictionary.columns where libname='WORK' and memname='SAMPLE_C'
          and varnum=1;
quit;
data sample_c2;
set sample_c;
Keep &firstvar_c  ;
run;

/*merge variable code and label*/
data sample_bc;
set sample_b2 (RENAME=(&firstvar_b=new_b));
set sample_c2 (RENAME=(&firstvar_c=new_c));
run;

/*create format format and output*/
data sample_bc;
set sample_bc;
Original_format = CATS(new_c,"=","'",new_b,"'");
run;

proc print data=sample_bc noobs;
   var Original_format;
run;

/*data _null_;*/
/*   set sample_bc ;*/
/*   file print;*/
/*   putlog Original_format @@;*/
/*   put Original_format @@;*/
/*run;*/

%mend formatd;


PROC FORMAT ;
    VALUE $GENDER
	  "M"   = "Male_new"
	  "F"   = "Female_new";
RUN;
DATA sample;
    SET sashelp.class;
    FORMAT sex $GENDER. ;
 RUN;

%formatd(sample, sex   );
```

### A macro to view the list of variables
```{r,engine='sas',error=TRUE}

%macro varnames (dat);
proc contents
     data = &dat
          noprint
          out = data_info
               (keep = name );
run;
proc print
     data = data_info
          noobs;
run;
%mend;
%varnames (sashelp.class) 
```


