
# SASmarkdown

## How to install sasmarkdown 
[sasmarkdown](https://www.ssc.wisc.edu/~hemken/SASworkshops/Markdown/SASmdpackage.html) 

 `Use an Engine`
```{r libload} 
# knitr::opts_chunk$set(echo = TRUE)
require(SASmarkdown)
```

```{r}
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -ls 75"
knitr::opts_chunk$set(engine='sashtml', engine.path=saspath,
                      engine.opts=sasopts, comment="")
```

## Common statements 
`Generally, can not use r syntax in these chunks ` 

`Data can not be used in different chunks`

- compute mean and freqency 
```{r , engine="sas" , error=TRUE}
proc means data=sashelp.class  ;
run;

proc freq data=sashelp.class  ;
table sex;
run;

proc contents data=sashelp.class varnum  ;
run;

```

- sort a data set
```{r , engine="sas" , error=TRUE}
proc sort data=sashelp.class Out= name ;
by name;
Run;

proc print data= name (obs=10) ;
run;
```

- transpose wide format data to long format data
```{r , engine="sas" , error=TRUE}
proc transpose data= sashelp.class out= Field;
by name;
Run;

proc print data= field (obs=10) ;
run;
```
- conditional statement
```{r , engine="sas" , error=TRUE}
 
DATA ab;
set sashelp.class;
IF sex="F" then message='A is greater';
Else message='B is greater';
Run; 

proc print data=ab (obs=10);
run;
```


 
- change format of the variable 
```{r , engine="sas" , error=TRUE}
proc format;
   value AGEnew
        11 = '1: NEW' 
        12 = '2: NEW'
        13 = '3: NEW' 
        14 = '4: NEW'
        15 = '5: NEW'
        16 = '6: NEW'        
        ;
        run;
        
DATA ab;
set sashelp.class;
Format AGE AGEnew.;
Run;
        
proc freq data=ab;
        table AGE;
        run;
        
```
- basic operations
```{r  , engine="sas" , error=TRUE}
Data Mathdata;
A= 10.12345;
B=20;
C= mean (a,b);
D= Min(a,b);
E= Max(a,b); 
F = log(a);
G= round(a,0.02);
H= floor(a );

Run;

proc print data=Mathdata;
run;

Data mathdata;
Set  sashelp.Iris;
Sum = sum (of SepalLength, SepalWIDTH);
Diff = SepalLength- SepalWIDTH;
Mult = SepalLength* SepalWIDTH;
Div= SepalLength/ SepalWIDTH;
Run;

proc print data=mathdata (obs=10);
run;
```
- rename a variable
```{r  , engine="sas" , error=TRUE}
 Data AB;
set sashelp.class; Rename AGE= AGENEW;
Run;

proc print data=AB;
run;
```
- extract text from a character  value
```{r  , engine="sas" , error=TRUE}
Data Mathdata;
Text = "Hello World";
Text1= substr(Text, 6, 2);
Run;

proc print data=mathdata  ;
run;
```
- convert the character value into a numeric value and reverse
```{r , engine="sas" , error=TRUE}
Data ABC;
set sashelp.class;
agenew= input (age, best.); 
Run;

proc print data=abc  ;
run;

```

```{r , engine="sas" , error=TRUE}
Data ABC;
set sashelp.class;
agenew= put (age, best.); 
Run;

proc contents data=abc  ;
run;
```

- change the length of the variable
```{r , engine="sas" , error=TRUE}
Data ABC;
set sashelp.class;
Length agenew $10.;
Label agenew=“New age”;
Run;

proc contents data=abc  ;
run;
 
```
- create a report
```{r , engine="sas" , error=TRUE}
proc report data=sashelp.class;
Column age;
Define age / display;
Run;
```

- create a random variable
```{r , engine="sas" , error=TRUE}
data ab;
set sashelp.class;
num=rand("normal");
run;

proc print data=ab (obs=10); 
run;
```
- combine two texts
```{r , engine="sas" , error=TRUE}
Data Mathdata;
Text = "Hello";
Text1= "World";
Text2= text || " " ||text1;
Run;

proc print data=Mathdata  ; 
run;
```
- compress spaces
```{r , engine="sas" , error=TRUE}
Data Mathdata;
Text = "Hello  World  ";
Text1= trim(text);
Text2= compress(text);
Run;

proc print data=Mathdata  ; 
run;
```

- identify the position of a specified text
```{r , engine="sas" , error=TRUE}
Data Mathdata;
Text = "Hello World";
indextext=  index(text, "or");
Run;

proc print data=Mathdata  ; 
run;
```
- convert upcase, lower case and propcase
```{r , engine="sas" , error=TRUE}
Data Mathdata;
Text = "Hello World";
upcase=  upcase(text );
lowcase=  lowcase(text );
propcase=  propcase(text );
Run;

proc print data=Mathdata  ; 
run;
```
- deduplication
```{r, engine="sas" , error=TRUE}
/* Dedup, original data has 19 observations  */
proc sort data = sashelp.class out =  dedup
nodupkeys;
by height; 
run;

proc print data= dedup;
run;

```
- select sub data set
```{r,engine="sas", error=TRUE}
data where;
set sashelp.class;
where sex ne "F";
/*if */
run;

proc print data= where;
run;
```


- create macro using do loop statement 
```{r,engine="sas",error=TRUE}
/*create a macro to calcualte descriptive stats */
%macro means(var_avg) ;

/*calculate means*/
proc means data=sashelp.class StackODSOutput n mean std  min p5 p95 max nmiss; 
var  &var_avg;
class sex;
ods output summary=result2;
run;

/*append then output*/
data masterresult2;                  * combine results;
set masterresult2 result2;
run;

%mend means;

/*use macro to merge all descriptive stats */
data masterresult2 ;
set _null_;
run;

%let vars= 
age
height
weight
;

%macro model ;
%do i=1 %to %sysfunc(countw(&vars));

%let x=%scan(&vars,&i);
 %means( &x )
 
%end;
%mend model;

%model;

proc print data= masterresult2;
run;

```
- output intermediate tables
```{r,engine="sas",error=TRUE}
ods   listing close;
ods trace on;
ods output ParameterEstimates= ParameterEstimates  ;
proc glm data=sashelp.class;
model height=age;
run;
ods trace off;
ods listing;

proc print data=ParameterEstimates;
run;

```
- merge data sets

`right, left join`
```{r,engine="sas",error=TRUE}
data class2;
input name $   score;
datalines;
Alfred 	85
Alice 89
Daniel 99
;
run;

proc sql;
create table class_c as
  select *
    from sashelp.class as a
	  left join class2 as b 
	    on a.name = b.name;
quit;

proc print data=class_c;
run;
	    
proc sql;
create table class_d as
  select *
    from class2 as a
	  right join sashelp.class as b 
	    on a.name = b.name;
quit;

proc print data=class_d;
run;

```
`full join`
```{r,engine="sas",error=TRUE}
data class2;
input name $   score;
datalines;
Alfred 	85
Alice 89
Daniel 99
;
run;
 
proc sql;
create table class_e as
  select *
    from sashelp.class as a
	  full   join class2 as b 
	    on a.name = b.name;
quit;

proc print data=class_e;
run;

```


`or using merge for appending`
```{r,engine="sas",error=TRUE}
data class2;
input name $   score;
datalines;
Alfred 	85
Alice 89
Daniel 99
;
run;
 
data class_f;
merge  sashelp.class class2;
by name;
run;

proc print data=class_f;
run;

```

`inner join`
```{r,engine="sas",error=TRUE}
data class2;
input name $   score;
datalines;
Alfred 	85
Alice 89
Daniel 99
;
run;
 
proc sql;
create table class_e as
  select *
    from sashelp.class as a
	  inner join class2 as b 
	    on a.name = b.name;
quit;

proc print data=class_e;
run;

```

`minus join`
```{r,engine="sas",error=TRUE}
data class2;
input name $   score;
datalines;
Alfred 	85
Alice 89
Daniel 99
;
run;
 
proc sql;
create table class_e as
  select *
    from sashelp.class as a
	  left join class2 as b 
	    on a.name = b.name 
	    where b.name is NULL;
quit;

proc print data=class_e;
run;

```




- create table 1
```{r , engine="sas" , error=TRUE}
  *** Load utility macros;
%include "C:\Users\hed2\Downloads\mybook2\mybook2\sasmacro\create_table1.sas";

  *** Specify input and output data sets, and the column variable.;
%let INPUT_DATA  = sashelp.class;
%let OUTPUT_DATA = Table1 ;
%let COLVAR      = sex;

  /*chort*/
 *** %AddText(text=Height);
 *** %CategoricalRowVar2(rowVar=);
%ContinuousRowVar2(rowVar=height );
 *** %AddText(text=);

%ContinuousRowVar2(rowVar=weight );
%ContinuousRowVar2(rowVar=Age );

proc print data= table1;
run;

/* Export Table 1 as a CSV file*/ 
proc export data=Table1 replace label
    outfile="C:\Users\hed2\Downloads\mybook2\mybook2\sasmacro\table1.csv"
    dbms=csv;
run;  

```
- create table 1 without statistical test
```{r , engine="sas" , error=TRUE}
 *** Load utility macros;
%include "C:\Users\hed2\Downloads\mybook2\mybook2\sasmacro/Utility Macros_updated_NG.sas";

/* 
  By  counts 
%let yourdata = sashelp.class;
%let output_data= By_sex;
%let formatsfolder= ; 
%let yourfolder = ;
%let varlist_cat =  ;
%let varlist_cont = age heigt weight; 
%let output_order = height weight age;
%let decimal_max =3;
%let group_by = sex;
%let group_by_missing=0;

%Table_summary;
*/
```


- perform a regression 
```{r , engine="sas" , error=TRUE}
proc glm data= sashelp.class;
model height=age;
Run;
```


