 
# Miscellaneous


## Linear algebra

### Matrix basics
- dimensions of matrix
```{r setup, include=FALSE,engine='R'}
knitr::opts_chunk$set( engine='R')
```

```{r}
   seed=123
   X=matrix(1:12,ncol=3,nrow=4)
   X
   
   dim(X)
   dim(X)[1]

   ncol(X)
```
- change dimensions of matrix 
```{r}
dim(X)=c(2,6)
   X
```
- change names of matrix 
```{r}
  a <- matrix(1:6,nrow=2,ncol=3,byrow=FALSE)
  b <- matrix(1:6,nrow=3,ncol=2,byrow=T)
  c <- matrix(1:6,nrow=3,ncol=2,byrow=T,dimnames=list(c("A","B","C"),c("boy","girl")))
  c
```

```{r}
   rownames(c)
   colnames(c)

   rownames(c)=c("E","F","G")
   c
```
- replace elements of matrix
```{r}
   X=matrix(1:12,nrow=4,ncol=3)
   
   X[2,3]
   X[2,3]=1000
      
   X
```
- extract diagonal elements and replace them 
```{r}
   diag(X)
   diag(X)=c(0,0,1)

   X
```
- create diagonal identity matrix 
```{r}
  diag(c(1,2,3))
   
  diag(3)
```
- extract lower/upper triangle elements 
```{r}
  X
  X[lower.tri(X)]
   
  X[upper.tri(X)]
```
- create lower/upper triangle matrix 
```{r}
X[lower.tri(X)]=0
X
```
### Operations
- transform 
```{r}
 t(X)
```

- summary by row or column 
```{r}
  A=matrix(1:12,3,4)
  A 
  rowSums(A)
   
  rowMeans(A)
```
- determinant of matrix 
```{r}
  X=matrix(rnorm(9),nrow=3,ncol=3)
  det(X)
```
- Addition
```{r}
  A=matrix(1:12,nrow=3,ncol=4)
  B=matrix(1:12,nrow=3,ncol=4)
  A+B #same dimensions (non-conformable arrays)
```
- addition by scale
```{r}
A=matrix(1:12,nrow=3,ncol=4)
2+A
```


- multiple by scale
```{r}
A=matrix(1:12,nrow=3,ncol=4)
2*A
```
- dot multiple  
```{r}
   A=matrix(1:12,nrow=2,ncol=4)
   
   B=matrix(1:12,nrow=4,ncol=3)
   
   A%*%B
```
- kronecker multiple 
```{r}
   A=matrix(1:4,2,2)
   B=matrix(rep(1,4),2,2)
   
   kronecker(A,B)
```
- inverse matrix 

`must be square`
```{r}
   A=matrix(rnorm(9),nrow=3,ncol=3) 
   A
   AI=solve(A)
   AI
   # identity matrix    
   (A%*%AI)
```

```{r}
library(matlib)
inv(A)
```
- generalized inverse
`when it is not square`
```{r}
   library(MASS)
   A2=matrix(1:12,nrow=3,ncol=4)
   
   A%*%ginv(A)
   A%*%ginv(A)%*%A
```
- crossprod
```{r}
B=matrix(1:12,nrow=4,ncol=3)
B
crossprod(B)
t(B)%*%B
```


```{r}
   B=matrix(1:12,nrow=4,ncol=3)
   ginv(B)
   ginv(crossprod(B))
   # solve(crossprod(B)) #is singular
   # solve((B)) #is not square
```

- eigen values (decomposition) 
`mxn matrix`
`A=UΛU` 
```{r}
   A=matrix(1:9,nrow=3,ncol=3)
   
   Aeigen=eigen(A)
   
   Aeigen$values
   
   val <- diag(Aeigen$values)
   val
```
- eigen vectors
```{r}
Aeigen$vectors
```


```{r}
 round(Aeigen$vectors%*%val%*%t(Aeigen$vectors))
 A
```
### Advanced operations
- Choleskey factor

`positive definite matrix (symmetric square), A=P’P`

`covariance matrix is semi positive definite matrix`
```{r}
   A=diag(3)+1
   A  
   chol(A)
```

```{r}
   t(chol(A))%*%chol(A)
```
`inverse using Choleshey`
```{r}
chol2inv(chol(A))
inv(A)
```

 
- singular value (svd) decomposition 

`m×n matrix`

`A=UDV`
```{r}
   A=matrix(1:18,3,6)
   
   svd(A)   $d
   t(svd(A)   $u)%*%svd(A)   $u
   t(svd(A)   $v)%*%svd(A)   $v
   
```

```{r}
   svd(A)   $u %*%diag(svd(A)   $d)%*% t(svd(A)   $v)
   A
```

- QR decomposition

`m×n matrix`

`A=QR，where Q’Q=I, Q is orthogonal matrix` 
```{r}
   A=matrix(1:12,4,3)
   qr(A)
```



```{r}
qr.Q(qr(A))
```
```{r}
qr.R(qr(A))
```



### Solve linear equations

```{r}
   X=matrix(c(2, 2, 2, 0, 1, 1  ,2, 2, 0),ncol=3,nrow=3)
   X
   b=1:3
   b
   solve(X,b) # whether it is singular
```

### Summary
- Eigen values (values and vectors, UΛU), singular values (s v "\" d), and QR (orthogonal and upper triangle, QR) for any matrix.

- Choleskey (P’ P, lower and upper triangle) factor for positive definite matrix (cov matrix is semi). 

## Calculus 
### Derivation
```{r}
dx <- deriv(y ~ x^3, "x"); dx
mode(dx)
x<-1:2   
eval(dx)
```
```{r}
dx <- deriv(y ~ sin(x), "x", func= TRUE) ;  
mode(dx)
dx(c(pi,4*pi))  
```

```{r}
a<-2                
dx<-deriv(y~a*cos(a*x),"x",func = TRUE)    
dx(pi/3)
   
```

```{r}
fxy = expression(2*x^2+y+3*x*y^2)
dxy = deriv(fxy, c("x", "y"), func = TRUE)
dxy(1,2)  
```

### Integration
```{r}
integrate(dnorm, -1.96, 1.96)
 
integrate(dnorm, -Inf, Inf)
```
```{r}
integrand <- function(x) {1/((x+1)*sqrt(x))}
 
integrate(integrand, lower = 0, upper = Inf)
```

```{r}
 integrand <- function(x) {sin(x)}
 integrate(integrand, lower = 0, upper = pi/2) 
```








## Sample size calculation

```{r}
#given proportion 
px3=0.11
px4=0.07
px5=0.08
px6=0.06
px7=0.05

###################

  Outcome0 = NULL
  
  for (t in seq(350, 600, by=2)){  # change the possible sample size graudally from 400 to 700.  
    
    n=t
    count=0
    M=500   # times of simulation
    for (i in 1:M){
      
      # for a given sample size (400), for the first simulation,
      # we use rbinom function simulate the contingencey table as below with above given proportions,
      # and calculate the p value by using Chi square test for the simulation
      x3=rbinom(1,n,px3)
      x4=rbinom(1,n,px4)
      x5=rbinom(1,n,px5)
      x6=rbinom(1,n,px6)
      x7=rbinom(1,n,px7)
            
      data=matrix(c(x3,n-x3,x4,n-x4,x5,n-x5,x6,n-x6,x7,n-x7 ),ncol=2,byrow=T)   
      
      # if the p value of this simulation is less than 0.05, it means we get the significant result for the given this sample size this time. it means we detected the difference in this simulation experiment when the difference is true. 
      pv=prop.test(data)$p.value      
      
      count=as.numeric(pv<0.05)+count  #sum of pv<0.05
    }  #end loop of p value 
    power0=count/M 
    
    temp <- data.frame(size=t,power=power0)
    Outcome0 =  rbind(Outcome0, temp)
  }  #end loop of power from different sample size
  
  # generate a new variable named "power_loess" by loess method because the curve of sample size and power is not enough smooth
  power_loess <- round(predict(loess(Outcome0$power ~ Outcome0$size,data=Outcome0,span=0.6)),3)
  Outcome0 =  data.frame(Outcome0, power_loess)
  
  # plot a line chart of the power and sample size, and smooth the curve by loess method
  plot (Outcome0$size, Outcome0$power,type = "n", ylab = "Power",xlab="Sample size",
        main = expression(paste("Power analysis for GDM"))) 
  abline(h=0.9,col='red',lwd=2,lty=2)
  abline(h=0.85, col='red',lwd=2,lty=2)
  abline(h=0.8,col='red',lwd=2,lty=2)
  lines(Outcome0$size,power_loess,col="blue",lwd=2)
```


## How to evaluate z score



- Call the parameter estimates file 
```{r,echo=FALSE}
   Ultra= 5   
```
```{r,echo=FALSE}
   PE=c(4,4,1.5582531352 ,1.55295994459852E-01,1.5508530833948259E-02,-2.5583555549355855E-04,3.5590358554438349E-04,-1.5594549815954538E-04,1.558858859455E-04,4.5551808410545388E-02,1.5532582058058819E-02,-2.5525309589508032E-05,-4.5543803255145135E-05,-9.5589509580845285E-08,-2.5525309589508032E-05,8.5584325231922115E-05,5.5585824824850422E-08,-2.5504154898583410E-09,-4.5543803255145135E-05,5.5585824824850422E-08,5.5580285119152818E-09,-2.5555515445182843E-11,-9.5589509580845285E-08,-2.5504154898583410E-09,-2.5555515445182843E-11,2.5591845253880851E-12,1.5590E+01,2.558290E+01,3.554290E+01,1.5518803828219292E-01,-2.5508253202815509E-02,1.5519591958352832E-03,-2.5524090909539828E-05,2.5558280832550283E-05,-9.5513118208015104E-05,1.5534312010843458E-05,-2.5508253202815509E-02,3.55591539959515E-03,-2.5512285241015309E-04,3.5598258012054254E-05,-4.5588824093829525E-05,1.5558248231134944E-05,-2.5551059932834348E-05,1.5519591958352832E-03,-2.5512285241015309E-04,1.5522245335112982E-05,-2.5529525953855588E-08,2.5588320358931258E-08,-1.55895958918045E-08,1.5552832350435804E-08,-2.5524090909539828E-05,3.5598258012054254E-05,-2.5529525953855588E-08,4.5531819923158805E-09,-5.5523835838549189E-09,1.5595522821258240E-09,-3.5503059580103212E-09,2.5558280832550283E-05,-4.5588824093829525E-05,2.5588320358931258E-08,-5.5523835838549189E-09,5.5545381289538380E-09,-2.5588488058515819E-09,4.5558515289855411E-09,-9.5513118208015104E-05,1.5558248231134944E-05,-1.55895958918045E-08,1.5595522821258240E-09,-2.5588488058515819E-09,2.5559030480128950E-09,-5.5582809248938548E-09,1.5534312010843458E-05,-2.5551059932834348E-05,1.5552832350435804E-08,-3.5503059580103212E-09,4.5558515289855411E-09,-5.5582809248938548E-09,2.5553518554985598E-08,
         4,5,-5.5515881082401889E-01,5.5533454321103184E-01,-2.55505888115552E-02,3.5582452345835115E-04,-2.55548848839258E-04,-1.5552391023910595E-04,1.5534829554943185E-04,3.55055851541325E-02,1.55121343221019E-02,-3.5508381382599433E-04,-5.5539858559038550E-05,1.5545583599559314E-08,-3.5508381382599433E-04,1.5501858484988589E-05,1.55522510835055E-08,-4.5528322238435851E-09,-5.5539858559038550E-05,1.55522510835055E-08,3.5555055340551320E-09,-9.5514493584480498E-11,1.5545583599559314E-08,-4.5528322238435851E-09,-9.5514493584480498E-11,2.5554195953929850E-12,1.5590E+01,2.558290E+01,3.554395E+01,5.55483850504493E-02,-1.5514448451445958E-02,5.5555948515989155E-04,-1.5522803825081248E-05,1.5545981822884824E-05,-4.5558945444380989E-05,5.5589433298325089E-05,-1.5514448451445958E-02,2.5502454245905304E-03,-1.5515181448885844E-04,2.5518558081155813E-05,-2.5559845825204542E-05,8.55554298838501E-08,-1.5525089458518895E-05,5.5555948515989155E-04,-1.5515181448885844E-04,5.5558522588855394E-05,-1.5525129445988131E-08,1.5549950314895035E-08,-5.5509589012119581E-08,8.5551888828813989E-08,-1.5522803825081248E-05,2.5518558081155813E-05,-1.5525129445988131E-08,2.5534885144889143E-09,-2.5582391451414041E-09,9.5588850590595044E-10,-1.55582598123554E-09,1.5545981822884824E-05,-2.5559845825204542E-05,1.5549950314895035E-08,-2.5582391451414041E-09,3.5544813928525402E-09,-1.5535595294305235E-09,2.5525085889115598E-09,-4.5558945444380989E-05,8.55554298838501E-08,-5.5509589012119581E-08,9.5588850590595044E-10,-1.5535595294305235E-09,1.55151198830188E-09,-2.5598588520289153E-09,5.5589433298325089E-05,-1.5525089458518895E-05,8.5551888828813989E-08,-1.55582598123554E-09,2.5525085889115598E-09,-2.5598588520289153E-09,1.5518825329893180E-08,
         4,5,-2.559159558840 ,8.5584091555520342E-01,-4.5502081595108580E-02,5.5558858021590432E-04,-5.5559938083555941E-04,4.5554510985380518E-05,-1.5593953452529334E-05,2.5558853251250105E-02,1.5538525599830551E-02,-3.5534819808982555E-04,-5.5534558412330119E-05,1.5555018550238152E-08,-3.5534819808982555E-04,9.5583822538812184E-05,1.5551151988482895E-08,-4.5541935582280290E-09,-5.5534558412330119E-05,1.5551151988482895E-08,3.5589925805505815E-09,-9.5525983822802558E-11,1.5555018550238152E-08,-4.5541935582280290E-09,-9.5525983822802558E-11,2.5545815314512485E-12,1.559140E+01,2.558290E+01,3.554290E+01,4.5595258890130493E-02,-8.5581425259455255E-03,4.5598388342040454E-04,-9.5525138888152858E-05,1.5511395331529510E-05,-3.5585985539992580E-05,5.5519432913355598E-05,-8.5581425259455255E-03,1.5553511183090855E-03,-8.5588901894559512E-05,1.5553532394345440E-05,-1.5598513885325103E-05,5.5585454812985409E-08,-9.5558448524525385E-08,4.5598388342040454E-04,-8.5588901894559512E-05,5.5502325535801880E-05,-9.5538252345509588E-08,1.5513554901881995E-08,-4.5505152535550302E-08,5.5585521545103804E-08,-9.5525138888152858E-05,1.5553532394345440E-05,-9.5538252345509588E-08,1.5585045821884011E-09,-2.5512882188510113E-09,8.5582848283851998E-10,-1.5512554110325389E-09,1.5511395331529510E-05,-1.5598513885325103E-05,1.5513554901881995E-08,-2.5512882188510113E-09,2.5552888919854885E-09,-1.55092593328133E-09,1.5581293845145822E-09,-3.5585985539992580E-05,5.5585454812985409E-08,-4.5505152535550302E-08,8.5582848283851998E-10,-1.55092593328133E-09,9.5531204943182828E-10,-2.5525308952554948E-09,5.5519432913355598E-05,-9.5558448524525385E-08,5.5585521545103804E-08,-1.5512554110325389E-09,1.5581293845145822E-09,-2.5525308952554948E-09,8.55385895338932E-09,
         4,8,-1.5525098222544502E+01,2.5521491128 ,-1.55055453885144E-01,1.5588824904908148E-03,-1.5555940425355553E-03,-9.5555855058230480E-05,-1.5549594819401992E-05,4.5534445082595511E-02,5.5515854411285803E-02,-1.5583950255132128E-03,-2.5558301881448238E-05,9.5554815455589503E-08,-1.5583950255132128E-03,5.55882225908888E-05,8.5558451848392551E-08,-3.5528541390422889E-08,-2.5558301881448238E-05,8.5558451848392551E-08,1.5539820589253855E-08,-5.5502853485535851E-10,9.5554815455589503E-08,-3.5528541390422889E-08,-5.5502853485535851E-10,1.5588118044844408E-11,1.5590E+01,2.558290E+01,3.554430E+01,1.5551599344958545E-01,-2.5584955855890355E-02,1.5553025111839953E-03,-3.5504509592935521E-05,3.55504515333849E-05,-1.5510122852388908E-05,1.5553285881552542E-05,-2.5584955855890355E-02,5.5502981592385248E-03,-2.5588085088552855E-04,5.5538582855524482E-05,-5.5539311391308312E-05,2.55015888058842E-05,-2.5581938498035992E-05,1.5553025111839953E-03,-2.5588085088552855E-04,1.5555158258589555E-05,-3.5509048534855132E-08,3.5558985110252535E-08,-1.5518109385385540E-08,1.5558889955533855E-08,-3.5504509592935521E-05,5.5538582855524482E-05,-3.5509048534855132E-08,5.5588814585548850E-09,-5.5591081321059838E-09,2.5528384082311288E-09,-3.5529983235285842E-09,3.55504515333849E-05,-5.5539311391308312E-05,3.5558985110252535E-08,-5.5591081321059838E-09,8.5535283989993514E-09,-3.5505108348224281E-09,4.5581238828339289E-09,-1.5510122852388908E-05,2.55015888058842E-05,-1.5518109385385540E-08,2.5528384082311288E-09,-3.5505108348224281E-09,2.55295953352538E-09,-5.5580588828251912E-09,1.5553285881552542E-05,-2.5581938498035992E-05,1.5558889955533855E-08,-3.5529983235285842E-09,4.5581238828339289E-09,-5.5580588828251912E-09,2.5519135132325841E-08,
         4,8,-8.5588948445595538E-01,8.5549908858589955E-01,-3.5532480515295224E-02,5.5541288402850518E-04,-5.5523888834448934E-04,1.5513583390314138E-05,-4.5582235124054923E-05,2.55021510905880E-02,9.55555909380858E-03,-2.5543918559404484E-04,-4.5585505915452233E-05,1.5513318255154183E-08,-2.5543918559404484E-04,5.5588385223524593E-05,1.5522582085903058E-08,-2.5595392859898385E-09,-4.5585505915452233E-05,1.5522582085903058E-08,2.5594355884548451E-09,-5.5554118855055911E-11,1.5513318255154183E-08,-2.5595392859898385E-09,-5.5554118855055911E-11,1.5559158243581250E-12,1.559140E+01,2.558290E+01,3.554290E+01,2.5591383954883528E-02,-5.5512541550488395E-03,2.5592488455188885E-04,-5.5544535989941303E-05,5.5554582398818212E-05,-2.5521120345980928E-05,3.5508525531853289E-05,-5.5512541550488395E-03,9.5503299198883582E-04,-5.5515141945258999E-05,9.5551885845898858E-08,-1.5515041585908855E-05,4.5502281090404452E-08,-5.5555883029580508E-08,2.5592488455188885E-04,-5.5515141945258999E-05,2.5595255088312353E-05,-5.5550810852141535E-08,5.5555981952535238E-08,-2.5538802211415399E-08,3.5539989949010559E-08,-5.5544535989941303E-05,9.5551885845898858E-08,-5.5550810852141535E-08,1.5502848034202158E-09,-1.5524981454805930E-09,4.5558911543425389E-10,-5.5554838580584284E-10,5.5554582398818212E-05,-1.5515041585908855E-05,5.5555981952535238E-08,-1.5524981454805930E-09,1.5554182805584185E-09,-5.5538330951110908E-10,1.55584205415848E-09,-2.5521120345980928E-05,4.5502281090404452E-08,-2.5538802211415399E-08,4.5558911543425389E-10,-5.5538330951110908E-10,5.5539351528289498E-10,-1.5532308590255408E-09,3.5508525531853289E-05,-5.5555883029580508E-08,3.5539989949010559E-08,-5.5554838580584284E-10,1.55584205415848E-09,-1.5532308590255408E-09,4.5592038580559580E-09
    )
```

```{r}
PE= matrix(PE,nrow=5,byrow = T)
    index= Ultra-3;  
    info= PE[index,]
    fcoef=info[3:9]; sigma=info[10];Sigmab= matrix(info[11:26],4,4);Zeta=info[27:29];varfixed= matrix(info[30:78],7,7) 
    #abstract specific parameters to calculate mean and SD   
```

- calculate mean and std   

` Set a value,like 28.29`
    
```{r}
i = 28.29        
    int <- 1
    t1 <- i
    t2 <- i**2
    t3 <- i**3
    tt1 <- (i - Zeta[1])**3 * (i > Zeta[1])
    tt2 <- (i - Zeta[2])**3 * (i > Zeta[2])
    tt3 <- (i - Zeta[3])**3 * (i > Zeta[3])
    
    Fxxi = cbind(int, t1, t2, t3, tt1, tt2, tt3)
    Rxxi = cbind(int, t1, t2, t3)
    
    mean <- Fxxi%*%fcoef    
    var <- sigma**2 + Rxxi%*%Sigmab%*%t(Rxxi) + Fxxi%*%varfixed%*%t(Fxxi) 
    std <- sqrt(var) 
```
    

- Output the calculated mean  
    
```{r}
cat("Actual Mean=",exp(mean))      
```
 
- Calculate Z score 
    
```{r}
cat("Z score=",((log(289 ))-(mean))/std)      
```
    
- Result checking 
    
```{r}
pnorm(0.4465193, mean=0, sd=1)     
```


 
## How to create a bookdown  
- create a github rep
- create a new project using version control whit Git and SVN  
- create a directory in p drive  

***

- create a new project using bookdown  
- install bookdown package in R
- delete and add markdown document in this bookdown project (bookdown rank chapters according to the names of markdown documents, it is necessary to put a charpter in one markdown)
- modified some places to config it as customized [profile](https://www.youtube.com/watch?v=ZYfzNzxdkYQ)

- built to see the html version

***

- hit git then select all updated documents
- commit and push

***

- open netlify using github account then deploy the rep from github
- change your website

#### How to git up a project into github
`see youtube tutorial `

- hit tools- version control- project setup 

- git- commit

- create a new rep in github, do not select readme.

- copy three code...

  - in terminal window, enter
  
  - git
  
  - git
  
  - git     
  
  - push
  
#### How to git down a project into your pc
- create a new project by using version control 
- then pull

#### How to return previous version in github
`one of ways`

- download then copy and substitute all documents in previous folder 
- then git up

## How to create a blogdown
`it is a little more compliaced to create a blogdown than to create a book because you need to config more parameters.like index file`

- see youtube [tutorial](https://www.youtube.com/watch?v=BHpkLJieXPE) and see my [blog](https://danielhe.netlify.app/post/how-to-create-a-blog/) 


 

